Evolutionary remodeling of non-canonical ORF translation in mammals

Version 1

Abstract

Non-canonical open reading frames (ncORFs) are pervasive within transcripts annotated as “non-coding” or “untranslated regions” of mRNAs, yet their landscape under normal physiological conditions remains to be fully resolved, particularly outside humans. Here we applied a stringent and standardized pipeline to hundreds of high-quality ribosome profiling libraries from normal mammalian tissues and cell types, identifying 11,623 human and 16,485 mouse ncORFs. Evolutionary analyses revealed that thousands of ncORFs are subject to coding constraint and exhibit lineage-specific conservation, underscoring their functional potential. Ancient ncORFs are preferentially highly translated, broadly expressed, and enriched for lineage-specific conservation. Co-expression patterns further indicate that many ncORFs, especially ancient ones, are cotranslated with canonical coding sequences, consistent with functions mediated through protein–protein interactions. Together, these findings establish a comprehensive atlas of mammalian ncORFs and provide fundamental insights into their evolutionary dynamics and functional integration within the proteome.

Introduction

The systematic identification of functional genomic elements represents a foundational objective in the field of genomics. Among these elements, non-canonical open reading frames (ncORFs)—which are open reading frames (ORFs) located within long non-coding RNAs (lncRNAs) or untranslated regions (UTRs) of mRNAs—have attracted considerable attention in recent years1–5. While certain ncORF subtypes, such as upstream ORFs (uORFs) located in 5’ UTRs and downstream ORFs (dORFs) in 3’ UTRs, have been extensively characterized for their ability to regulate mRNA translation and stability in cis, a growing body of evidence demonstrates that ncORFs can also give rise to functional proteins that engage in diverse cellular pathways1,2,6,7. As functional ncORFs continue to be discovered across multiple eukaryotes8–17, systematic elucidation of their composition and function promises to uncover an additional layer of regulatory complexity and to refine our conception of the so-called “non-coding” genome.

Although ncORFs may play important biological roles, the number of translated ncORFs in the human genome remains uncertain. Detection by mass spectrometry (MS)-based proteomics is challenging, likely due to their short lengths and low abundances18. Ribosome profiling (Ribo-Seq) offers greater sensitivity for identifying translation events18,19, but downstream computational pipelines often produce highly divergent ncORF sets 20. Consequently, published estimates of ncORF abundance span several orders of magnitude—ranging from thousands to millions 15–22. To establish a standardized reference, the GENCODE consortium integrated data from seven distinct studies to produce a catalog of 7,264 human ncORFs (hereafter GENCODE ncORFs)2; however, the heterogeneity of sample sources and prediction algorithms likely introduced systematic bias. This catalog is also restricted to AUG-initiating ORFs, thereby overlooking those with near-cognate start codons. A later effort generated a high-resolution human translatome map and identified 7,767 ncORFs with a unified pipeline22. Nevertheless, this conservative annotation disproportionately under-represents coding sequence (CDS)-overlapping ncORFs. Indeed, later estimation suggests that between 10,500 to 22,500 ncORFs should exist in the human genome18. Furthermore, many ncORFs were reported from studies of cancer cells1,13,23,24, where extensive transcriptome reprogramming alters the translational landscape25,26. While paired healthy tissues were also analyzed in these studies, the landscape of ncORFs translated under normal physiological conditions is not yet fully resolved.

While a limited subset of ncORFs has been shown to encode signaling molecules, cofactors, membrane proteins, or subunits of large protein complexes1,27,28, functions of most ncORF-encoded proteins (ncEPs) are largely uncharacterized. Advances in high-throughput screen technologies have enabled the systematic interrogation of functional ncEPs23,28,30–33. Nevertheless, there is minimal concordance among hits from different studies3. Evolutionary analyses can offer insights into potential functionality of these translation products and mechanisms of origination34–36, yet how ncORF translation has been remodeled to integrate into proteome since their evolutionary origins is still poorly understood. Consequently, the extent to which translated ncORFs encode functional proteins and the evolutionary dynamics of their integration into the existing proteome are largely unresolved. Moreover, because most ncORF studies have focused on humans, it is unclear whether ncORFs in other species share similar sequence features and expression patterns.

To address these gaps, we generated a standardized annotation of translated ncORFs by uniformly reprocessing ∼400 high-quality ribosome profiling libraries from normal mammalian tissues and cell types. This stringent approach yielded 11,623 human and 16,485 mouse ncORFs, consistently recovered across diverse samples and prediction strategies. Comparative and functional genomic analyses revealed that thousands of these ncORFs are under evolutionary constraint and exhibit distinct translation profiles. The resulting dataset provides a robust, comprehensive resource that will facilitate future investigations into the biological roles of ncORFs across cellular and organismal contexts.

Results

Annotation of ncORFs translated in normal mammalian cells

We developed a standardized pipeline to systematically annotate ncORFs translated in normal human and mouse cells (Fig. 1A). We first curated publicly available Ribo-Seq libraries, excluding those from pathological or genetically modified samples. Since both library size and the proportion of in-frame ribosome-protected fragments (RPFs) affect the sensitivity and accuracy of ncORF prediction20, we filtered out libraries containing fewer than 5 million RPFs or with in-frame rates below 60%. After this quality control step, we retained 174 human and 209 mouse high-quality Ribo-Seq libraries, spanning a range of tissues and cell types (Fig. S1; Table S1). We predicted translated ORFs that begin with NUG (where N stands for any nucleotide) start codons and encode at least five amino acids using three tools— PRICE37, RiboCode38 and Ribo-TISH39—which we previously benchmarked for optimal performance20. After excluding annotated CDSs and their extension or truncation variants arising from alternative initiation sites, we obtained a preliminary list of ncORFs in each library.

Annotation of ncORFs in humans and mice.

(A) Overview of the integrative pipeline for ncORF annotation in human and mouse genomes. (B-C) RPF coverage plot for a representative uORF (B) and a lncORF (C), visualized using the Ribo-Seq signal track from GWIPS-viz68. ncORFs were shown in yellow and CDSs in cyan. Dashed lines indicate identical genomic regions shared across transcript isoforms. (D-E) Proportion of GENCODE ncORFs rediscovered in this study. GENCODE ncORFs were stratified either by reproducibility across independent studies (D) or by tiered translation evidence strength (E). (F) Known ncORFs that were experimentally characterized in earlier studies and independently rediscovered in this study from humans and mice. CDS, coding sequence; FLOSS, fragment length organization similarity score; MANE, matched annotation from NCBI and EBI; ncORF, non-canonical open reading frame; ORF, open reading frame; RPF, ribosome-protected fragment.

To generate reliable ncORF annotations, we applied stringent filtering and integration criteria. We removed potential false positives caused by non-ribosomal contamination or translational noise based on metrics including Fragment Length Organization Similarity Score (FLOSS)40, Ribosomal Release Score (RRS)41,42 and the number of RPFs within an ORF by P-site. We then aggregated the remaining ORFs to identify those consistently detected across multiple methods and libraries in human and mouse datasets, respectively (Fig. 1A). In many cases, two ORFs predicted from different libraries or methods may share the same start or stop codon due to the alternative translation initiation or splicing, resulting in groups of overlapping ncORFs. For example, 11.1% (805) of GENCODE ncORFs overlap with at least one other ORF, forming 329 overlapping groups (Table S2). This overlap complicates the identification of reproducible ORFs, as ORFs within a group may represent distinct proteoforms of a protein. To address this issue, we collapsed each group of ncORFs sharing the same start or stop codon into a single cluster, and retained only those clusters supported by at least two independent methods and observed in more than one library. For each retained cluster, we selected a representative ORF by prioritizing transcript annotation43,44, start codon type, and ORF length to facilitate downstream analyses (Fig. 1A).

This approach yielded 12,469 translated ncORFs in humans and 16,960 in mice. However, manual inspection revealed a subset of spurious ncORFs that are not truncation or extension variants of annotated CDSs, but instead partially overlap with annotated CDSs from alternative transcript isoforms in frame (Figs. 1B & C). For instance, in gene AGTPBP1, the CDS of transcript ENST00000357081 begins downstream of ENST00000628899 CDS due to the inclusion of an alternative exon. As a result, the 5’ UTR of ENST00000357081 appears to contain a translated uORF (Fig. 1B), which overlaps in-frame with the CDS of ENST00000628899 and may be mistakenly identified as a translated ncORF owing to active translation of the latter. A similar case is seen with the lncRNA ORF (lncORF) in ENST00000462211, which overlaps in-frame with the CDS of ENST00000328767 from the TMTC4 gene and may thus be erroneously identified as a translated ncORF (Fig. 1C). To avoid such artifacts, we implemented an additional filtering step to exclude these overlapping cases and left with 11,623 and 16,485 high-confidence ncORFs in humans and mice, respectively (Table S3).

Annotated ncORFs are supported by independent evidence

To assess whether the annotated ncORFs are supported by independent evidence, we compared them with a previously compiled catalog of ncORFs identified using mass spectrometry (MS)-based proteomic data20. For the 3,494 MS-supported ncORFs in humans and 873 in mice, 44.5% and 67.6%, respectively, were also predicted as translated ncORFs in this study (Fig. S2). We further evaluated our annotation against the GENCODE ncORFs2, a widely used community resource. Of the 7,264 GENCODE ncORFs, 6,974 are located in fully annotated transcripts and were included in our analysis. Among them, 2,997 were reported by at least two independent studies (hereafter “phase 1”), while the remaining were only reported in a single study (“single”)2. We found that 75.6% of phase 1 ncORFs were independently recovered in our annotation, significantly more than the single-study ncORFs (Fig. 1D; P = 6.4×10-307, Fisher’s exact test).

A recent study conducted a comprehensive pan-proteome analysis to provide high-confidence peptide evidence for GENCODE ncORFs45. In this study, ncORFs were categorized into four tiers: tier 1 and 2 ncORFs had support from both MS and Ribo-Seq data; tier 3 had MS evidence only; and tier 4 had Ribo-Seq evidence only29. Notably, our annotation was significantly enriched for tier 1 or 2 ncORFs (P = 2.0×10-64, Fisher’s exact test), with 446 and 661 of them detected, respectively (Fig. 1E). Combining MS-supported ncORFs from our previous compilation and those from GENCODE ncORFs, 2,088 (18.0%) human ncORFs identified in this study were supported by MS evidence. Despite this overlap, the majority (65.2%) of human ncORFs in our annotation are not included in either of the two catalogs. This highlights both the complexity of the non-canonical translatome and the critical role of tissue diversity in comprehensive ncORF annotation.

To further validate our findings, we manually curated ncORFs with known functions from the literature. We identified 17 ncORFs in our annotation that have been experimentally confirmed to be translated—either through reporter assays or antibody detection—and shown to produce functional proteins (Fig. 1F; Table S4). Notably, 16 of these validated ncORFs are from humans, and only 12 are included in the GENCODE catalog. This analysis further underscores the accuracy and functional relevance of our annotations. Together, these results indicate that our annotation pipeline generates a robust and reliable catalog of translated ncORFs in mammals.

ncORFs exhibit signatures of translation and encode putative proteins lacking known domains

We classified ORFs into categories based on community recommendations2 and found that the distribution of ORF types closely aligns with previous estimates18. In both human and mouse, uORFs, upstream overlapping ORFs (uoORFs), and lncORFs are more prevalent than other types, whereas internal ORFs (iORFs), downstream overlapping ORFs (doORFs), and dORFs are relatively rare (Fig. 2A). Although most ncORFs are short, encoding fewer than 100 amino acids (Fig. 2B), their sequences show clear signatures of translation similar to CDSs. These include preferential usage of AUG start codons (Fig. S3A), enrichment toward 5’ transcript ends (Fig. S3B), and reduced RNA secondary structure46 (Fig. S3C) and optimized Kozak contexts47,48 around start codons, especially in non-AUG ncORFs (Figs. S3D & S4). Despite these similarities, ncORFs differ markedly from CDSs in both codon and amino acid usage (Figs. S5A-B). This divergence is further supported by their lower tRNA adaptation index values and higher effective number of codons, indicating less optimized codon usage (Figs. S5C-D). Together, these findings show that ncORFs share core sequence and structural features with canonical CDSs, yet maintain distinct codon and amino acid usage profiles.

Sequence features of ncORFs and putative ncEPs in humans and mice.

(A) Number of ncORFs belonging to different categories. (B) Distribution of ncORF lengths. (C) Proportion of ncEPs that contain known Pfam domains. (D) Number of ncORFs with Pfam domains among lncORFs and other ncORF categories. (E) Distribution of the proportion of intrinsically disordered residues in CDS- and ncORF-encoded proteins. (F) Most frequently identified Pfam domains among human and mouse ncORFs. (G) Proportion of ncORFs overlapping with TEs, stratified by presence or absence of Pfam domains. Differences were assessed using Wilcoxon rank-sum tests. CDS, coding sequence; ncORF, non-canonical open reading frame; ncEP, ncORF-encoded protein; TE, transposable element.

Although ncORFs can encode putative proteins, the functional significance of ncEPs remains poorly understood. To assess their potential functions, we searched for known protein domains within their sequences. Overall, only 1.29% (150) of human and 0.78% (129) of mouse ncEPs contained at least one known domain (Fig. 2C; Table S5), presumably due to their short lengths relative to canonical proteins. Consistently, domain-containing ncORFs are significantly longer than those without annotated domains (Fig. S6) and are predominantly classified as lncORFs (Fig. 2D). Unlike canonical proteins, which have relatively low intrinsically disordered region (IDR) content (median ± standard error: 14.6±0.28% in humans and 12.7±0.26% in mice), ncEPs exhibit significantly higher IDR proportions (Fig. 2E), with medians of 62.5±1.42% and 60.0±1.27%, respectively. These results indicate that most ncEPs lack structured domains and are instead dominated by IDRs.

Among the eight domains observed in at least five ncORFs, three are DNA-binding domains: zinc finger, Krüppel-associated box, and helix-turn-helix (Fig. 2F). We hypothesize that these domains might originate from transposable elements (TEs), which often encode DNA-binding proteins involved in transposition49. Indeed, domain-containing ncORFs were significantly enriched for TE overlap in both humans and mice (P = 5.8×10-4 and 3.4×10-17, respectively; Fisher’s exact tests; Fig. 2G). Together, these results suggest that only a subset of longer ncORFs have acquired modular domains, potentially through TE-related mechanisms.

Putative ncEPs are selectively constrained

Given the scarcity of known domains in putative ncEPs, a key question is whether they are functional in cells. If at least some of these ncEPs are not merely translational noise or byproducts of translation-dependent functions, we would expect them to exhibit evolutionary signatures of selective constraints similar to canonical CDSs. To test this, we leveraged the recent high-resolution genomic constraint maps derived from both intra-and inter-species comparisons in humans 50,51. First, we examined the genomic non-coding constraint of haploinsufficient variation (Gnocchi), which quantifies selective pressure by comparing observed versus expected variants across human genomes. A Gnocchi score above zero indicates purifying selection50. Although Gnocchi scores of ncORFs are lower than those of CDSs, they are significantly greater than zero (Fig. S7), suggesting that ncORF sequences are under evolutionary constraint. A hallmark of functional coding regions is codon position-specific constraints, but Gnocchi scores lack the resolution needed to assess such fine-scale features. Therefore, we next analyzed the base-resolution conservation metric PhyloP52 calculated from multiple alignments of primate or mammalian genomes51. As expected, PhyloP scores in CDSs are substantially higher than those in flanking UTRs and display strong three-nucleotide periodicity, with the third codon position being less conserved than first and second (Fig. 3A). Although ncORFs (excluding those overlapping annotated CDSs) do not show higher overall PhyloP than flanking regions, they also exhibit significant three-nucleotide periodicity in both mammals (Fig. 3A) and primates (Fig. S8), suggesting that at least a subset of ncORFs are subject to similar evolutionary constraints as CDSs.

Evolutionary constraints of ncORFs.

(A) The mean PhyloP scores of 30 base pairs upstream and downstream of the ORF start or stop codons in mammals. Codons were delineated with bars of alternating colors by different frames, and nucleotides in untranslated regions were shown in grey. The significance of three-nucleotide periodicity was assessed by autocorrelation with a lag of three. (B) Cladograms of vertebrates with the number of ncORFs originating at each ancestral branch of humans. Triangles indicate species merged into larger clade for visual simplicity. (C) Relationship between ORF origination rates and node ages measured as node-to-tip distances. Origination rate was defined as the number of ncORFs that originated on a branch divided by the branch length. Blue lines show linear regression fits, and grey bands represent 95% prediction confidence intervals. Spearman’s correlation is indicated. (D) Distribution of ncORF PhyloCSF scores normalized by the number of codons in per ncORF. The dashed line denotes zero. CDS, coding sequence; ncORF, non-canonical open reading frame; ORF, open reading frame.

Motivated by the above results, we next sought to identify individual ncORFs under selective pressure. To this end, we determined their modes of origination34,35 and evolutionary history through comparative genomic analyses. Our analysis revealed that most ncORFs are evolutionarily young, with 63.0% (7,319) and 82.8% (13,508) ncORFs in humans and mice originating within the mammalian lineage, respectively (Figs. 3B & S9). In line with previous studies34,35, we estimate that at least 51.7% of human and 50.8% of mouse ncORFs arise de novo (Table S6). We also observed that the number of ncORFs originating at each ancestral node, when normalized by branch length, decreases with increasing node-to-tip distance (Fig. 3C), reflecting dynamic turnover and reduced long-term retention of ncORFs. To assess whether individual ncORFs are constrained for coding, we analyzed codon substitution patterns within ncORFs since origination34 with PhyloCSF53. PhyloCSF estimates the log-likelihood ratio of a given alignment under coding versus non-coding models; positive values indicate stronger similarity to conserved coding sequences. We found 13.8% (1,608) of human and 32.0% (5,281) of mouse ncORFs have positive PhyloCSF scores (Fig. 3D), and 10.9% (1,264) of human and 22.6% (3,721) of mouse ncORFs even have PhyloCSF exceeding 10 (Table S6), providing strong evidence that these sequences may encode functional ncEPs.

Evolutionary dynamics can also provide insights into the function of ncORFs, as functionally constrained ncORFs are less likely to be lost over time. To quantify the conservation of each ncORF since its origination, we developed a local branch length score (BLS) (Fig. 4A), adapting the “global” BLS metric previously used to assess uORF conservation across full phylogenies48,54. Unlike global BLS, local BLS is designed to detect lineage-specific constraint, making it better suited for identifying functional ncORFs that may be restricted to certain evolutionary branches. Using a local BLS threshold of 0.9, we identified 16.3% (1,889) of human and 24.8% (4,087) of mouse ncORFs as exhibiting lineage-specific conservation (Figs. 4B & S9). Notably, these ncORFs are significantly more likely to show positive PhyloCSF scores (Fig. 4C), reflecting the effectiveness of local BLS. In total, we identified 681 human and 1,622 mouse ncORFs that are both lineage-specifically conserved and show evidence of coding potential, making them strong candidates for future functional validation. Together, these findings suggest that a subset of ncORFs may encode functional proteins rather than translational byproducts.

Lineage-specifically conserved ncORFs.

(A) Schematic illustration for BLS calculation. (B) Cladograms of vertebrates showing the number of lineage-specifically conserved ncORFs (local BLS > 0.9) at each ancestral node for humans and mice. (C) Distribution of PhyloCSF scores per codon for lineage-specifically conserved ncORFs compared to all other ncORFs. BLS, branch length score; ncORF, non-canonical open reading frame; ORF, open reading frame.

Expression of ncORFs is shaped by their evolutionary dynamics

Gene expression is a tightly regulated process, and non-functional products are unlikely to be robustly expressed. To assess the expression of ncORFs, we examined their translation across diverse tissues and cell lines (Table S7). On average, 71.9% of human and 48.5% of mouse ncORFs are translated with reads per kilobase per million mapped reads (RPKM) ≥ 1, which is comparable to or slightly exceeds the proportions observed for CDSs (Fig. S10). To investigate the influence of ncORF evolution on their expression, we analyzed mean translation levels of ncORFs across different samples. We found that ncORFs with higher local BLS tend to be more highly translated in both species (Fig. S11A), whereas coding potential showed only a minor influence (Fig. S11B). To disentangle the effects of evolutionary constraint and gene age, we stratified ncORFs by origination nodes and examined their translation levels in relation to local BLS and PhyloCSF. Although ancient ncORFs are generally more highly translated, those with higher local BLS scores consistently show elevated translation levels compared to their age-matched counterparts (Figs. 5A & S12A). This suggests that lineage-specific conservation is associated with enhanced and possibly optimized translation, potentially reflecting acquired functional roles. Conversely, the relationship between coding potential and translation is more nuanced: young ncORFs with positive PhyloCSF scores tend to be more highly translated, while the opposite trend is observed in ancient ncORFs (Figs. 5B & S12B). These patterns point to complex interplays between evolutionary constraint, coding potential, and translational output, likely reflecting divergent selective pressures across different ncORF age groups.

Evolutionary dynamics of ncORF expression.

(A) Distribution of mean translation levels of human ncORFs grouped by their origin nodes and further stratified by local BLS. Statistical significance was assessed using Wilcoxon rank-sum tests. ***, P < 0.001; **, P < 0.01; *, P < 0.05. (B) Similar to (A) but ncORFs are further stratified by local BLS. (C) Relationship between ncORF tissue specificity at translation and transcription levels. Differences were determined with Wilcoxon signed-rank tests. (D-E) Similar to (A) and (B) but showing the distribution of tissue specificity at the translation level. CDS, coding sequence; ncORF, non-canonical open reading frame; ORF, open reading frame.

While the above analyses revealed the overall trends in ncORF translation, they overlooked the heterogeneity across tissues and cell types. We found that ncORF translation exhibits strong tissue specificity (Fig. S13), which may stem from either transcriptional bias in ncORF-containing genes or post-transcriptional regulation specific to ncORFs. To disentangle these factors, we examined RNA levels of ncORF-containing genes across different tissues55,56. Although transcriptional enrichment in testis was evident in both humans and mice (Fig. S14), testis-biased translation was observed only in mice (Fig. S13). Moreover, ncORF translation displayed significantly higher tissue specificity than the transcription of their host genes in both species (Fig. 5C), indicating a substantial post-transcriptional component to their regulation. We next assessed how evolutionary characteristics shape tissue specificity. Ancient ncORFs tend to have lower tissue specificity than younger ones, suggesting they are more broadly translated (Figs. 5D & S12C). Intriguingly, coding potential exhibits opposite effects depending on evolutionary age: young ncORFs with higher coding potential are more broadly expressed, whereas ancient ones show increased tissue specificity (Figs. 5E & S12D). These patterns may help to explain the lower average translation levels observed for ancient ncORFs across tissues. A similar trend was also evident for ncORFs with high local constraint scores in mice (Fig. S12C). Collectively, these results support a model in which ncORFs initially evolve widespread expression to establish interactions and avoid loss57, but may later become functionally specialized, resulting in more restricted expression.

Extensive co-translation between ncORFs and canonical CDSs

Given our earlier observation that many ncEPs lack modular domains and may act via interactions with canonical proteins, we next explored the landscape of co-translation between ncORFs and CDSs across different tissues or cell types as we expect that interacting proteins tend to be translated simultaneously spatiotemporally. Using weighted gene co-expression network analysis (WGCNA)58, we identified 61,513 human and 55,449 mouse ncORF-CDS co-translated pairs. As expected, ncORFs in mRNAs tend to co-translate with the main CDS of the same gene, even after excluding CDS-overlapping ones (Fig. S15), consistent with previous experimental findings28. This co-translation forms a highly modular bipartite network, in which one CDS may be linked to multiple ncORFs and vice versa (Fig. 6A-B). Network analysis revealed that hub CDS nodes—those with the most connections—in the largest modules are significantly enriched for functional gene categories (Fig. 6C-D). For instance, the largest module in humans is associated with chromatin-related proteins, while the second largest module in humans and the largest one in mice are linked to extracellular matrix organization. Notably, ancient ncORFs (those originating before the last common ancestor of mammals) are significantly more likely to co-translate with CDSs than younger ones (Fig. 6E), potentially reflecting a greater likelihood of acquiring functional interactions over evolutionary time. In summary, our analyses revealed that evolutionary dynamics of ncORFs have widespread impacts on their translation and potential function.

ncORF-CDS co-translation network.

(A) Bipartite network of ncORF-CDS co-translation in humans. (B) Bipartite network of ncORF-CDS co-translation in mice. ncORFs and CDSs are represented by different shapes and colored according to their cluster membership. Only the largest clusters were highlighted (top two in humans and top five in mice). (C) Top five enriched gene ontology terms for each of the two largest clusters in the human network shown in (A). (D) Top five enriched gene ontology terms for each of the three largest clusters in the mouse network shown in (B). (E) Proportion of ncORFs cotranslating with CDSs among ancient (pre-mammalian origin) versus younger (mammalian-specific) ncORFs. Differences were tested with Fisher’s exact tests. CDS, coding sequence; ncORF, non-canonical open reading frame; ORF, open reading frame.

Discussion

Reliable annotation of translated ncORFs is critical for understanding their functional roles and evolutionary significance. In this study, we developed a standardized pipeline for the annotation of translated ncORFs and applied it to both humans and mice due to the extensive publicly available Ribo-Seq datasets. Compared to previous efforts, our annotation framework offers several key improvements. First, we selected only high-quality libraries with sufficient read coverage and acceptable in-frame RPF rates. Our earlier work demonstrated that both library size and in-frame rates impact the reproducibility and accuracy of ncORF predictions20. This step helps mitigate false annotations arising from suboptimal data quality. Second, although many previously characterized functional ncEPs are associated with cancer-related ncORFs1, ncORFs may also have important functions under physiological conditions. Importantly, such ncORFs are more likely to be subject to evolutionary constraints. Therefore, we restricted our analysis to normal cells and tissues, allowing us to identify more biologically relevant and conserved ncORFs. Third, given the variable performance across ncORF detection methods and the low reproducibility of predictions between biological replicates20, we implemented stringent reproducibility criteria. Specifically, we retained only ncORFs detected by multiple methods and across independent datasets, enhancing the robustness of our final annotation. Finally, we applied a suite of well-validated filtering metrics from prior studies40–42 to minimize the inclusion of false positives, including those arising from translational noise59,60 or non-ribosomal contamination. As a result, our annotated ncORFs show significant enrichment for GENCODE ncORFs, especially those with MS evidence or reproducible detection across independent studies. These features collectively underscore the reliability and utility of our annotation for future functional and evolutionary studies of non-canonical translation.

From the perspective of new gene evolution, ncORFs represent a transitional stage between non-coding DNA/RNA and protein-coding genes61,62. A major challenge in reclassifying ncORFs as bona fide coding genes lies in determining whether the putative proteins they encode are functionally relevant. Our analysis of ncEPs revealed that most lack recognizable protein domains and are enriched for IDRs. Therefore, we propose that many ncORFs may function in a non-autonomous manner by interacting with other proteins to exert their effects. This hypothesis aligns with prior studies showing physical interactions between ncEPs and known proteins28,30,35. Supporting this view, our WGCNA analysis uncovered widespread co-translation between ncORFs and canonical CDSs, suggesting potential interaction partners and functional contexts.

We also assessed whether ncORFs are subject to evolutionary constraints consistent with functional protein coding. Using both population-level and cross-species analyses, we found that many ncORFs are under purifying selection, and their sequence conservation patterns often exhibit base-level features reminiscent of canonical CDSs—albeit to a lesser degree. Furthermore, we reconstructed the evolutionary trajectories of thousands of ncORFs in both humans and mice and observed sustained coding potential since their origination. Complementing these findings, we identified a subset of ncORFs that preserve intact ORF structures across most descendant species, suggesting ongoing purifying selection against disabling mutations. These lineage-specifically conserved and potentially coding ncORFs likely contribute to clade-specific functional innovation and represent compelling candidates for future functional characterization.

We further explored how evolutionary history shapes ncORF translation across tissues or cell types. Our results indicate that ancient ncORFs are more broadly and highly translated and have established more extensive co-expression with canonical CDSs. Moreover, when comparing ncORFs of similar age, those that are lineage-specifically constrained consistently exhibit higher translation levels, reinforcing their likely functional importance. Interestingly, we found that young ncORFs with strong evolutionary constraints tend to have lower tissue specificity in their translation, potentially reflecting a phase of widespread expression that facilitates functional recruitment. In contrast, older ncORFs with positive coding potential tend to display increased tissue specificity, implying functional specialization over evolutionary time. Collectively, these findings support a model in which ncORFs evolve from broadly expressed, weakly functional elements into specialized, conserved effectors— though further empirical work is needed to validate this hypothesis.

In conclusion, we provide a comprehensive annotation of translated ncORFs in humans and mice, offering a valuable resource for functional and evolutionary genomics. Our integrated analysis of sequence evolution and translational landscape sheds light on the evolutionary trajectory and functional integration of ncORFs. The consistency of these patterns in both species suggests that our findings may generalize to other mammals. However, realizing the full biological significance of ncORFs, particularly lineage-specific ones, will require experimental validation. Recent advances in high-throughput screening technologies19,30,33,63 offer a promising path forward. Lastly, although our study focused on humans and mice due to the availability of high-quality Ribo-Seq data, we anticipate that our annotation will serve as a training resource for machine learning models aimed at predicting translated ncORFs in non-model organisms, thus broadening our understanding of genome coding potential across species.

Methods

Quality control and preprocessing of Ribo-Seq libraries

Ribo-Seq libraries from normal human or mouse samples without genetic modifications or perturbations (Table S7) were downloaded from the European Nucleotide Archive (ENA) or the Short Read Archive (SRA). For ncORF annotation, libraries containing fewer than five million reads were excluded. Libraries derived from cancer tissues or cell lines were also removed based on sample metadata or the associated publications. Next, adapter sequences and any random nucleotide additions were trimmed using Cutadapt (v3.7)64 and reads shorter than 18 nucleotides were discarded. Remaining reads were aligned to the human or mouse reference genome (GRCh38 or GRCm39, ENSEMBL65 release 107) using the STAR aligner (v2.7.10a)66. To assess library quality, the proportion of in-frame ribosome-protected fragment (RPF) reads was calculated using the PSite package67. Libraries with ≥60% in-frame RPFs were classified as high quality. Only samples containing at least two high-quality libraries were retained for ncORF annotation (Table S1). For the expression analyses, all the downloaded Ribo-Seq libraries were used to cover more samples.

ncORF annotation

For each Ribo-Seq library, translated ORFs initiating with AUG, CUG, GUG, or UUG and encoding at least five amino acids were predicted using PRICE37, RiboCode38, and Ribo-TISH39, as described previously20. To define novel ORFs, annotated CDSs and their truncation or extension variants (i.e., ORFs with the same start or stop position as annotated CDSs) were excluded. To eliminate likely false positives resulting from non-ribosomal contamination or stochastic molecular errors, we calculated RRS41,42 and FLOSS40 of each ncORF. RRS measures the drop in RPF coverage after stop codon—a key signature of active translation beyond three-nucleotide periodicity41,42, whereas FLOSS assesses whether the RPF length distribution resembles that of canonical CDSs, helping to distinguish authentic ORFs from non-ribosomal signals40. In each library, uniquely mapped RPFs were assigned to the corresponding P-site, and genome-wide P-site coverage was computed with the PSite package. Similar to previous studies41,42, RRS was calculated as the sum of coverage across the five codons preceding the stop codon divided by the sum across the five trinucleotides following the stop codon. FLOSS was calculated by first deriving a reference RPF length distribution vector (Pref) using all RPFs uniquely mapped to annotated CDSs. For each ORF (ncORF or CDS), a read length frequency vector (PORF) was computed, and FLOSS was calculated as ∑|Pref - PORF|/2. The empirical upper bound for acceptable FLOSS was set using Tukey’s method (i.e., the third quartile + 1.5 × inter-quartile range of FLOSS values across all CDSs)40. ncORFs with fewer than 10 P-site-mapped RPFs, RRS ≤ 1, or FLOSS values exceeding the empirical upper bound were filtered out.

Integration of ncORF prediction across different libraries of the same species was nontrivial due to extensive overlap among ncORFs in genomic coordinates. To address this issue, we used a graph-based clustering approach. We created a graph where two ORF vertices were connected if they share the same start or stop codon. Then, each set of connected ncORFs in the graph represents a cluster of overlapping ncORFs. To ensure reproducibility, only clusters identified by ≥2 methods in ≥2 libraries were retained, yielding 13,071 clusters in humans and 17,674 clusters in mice. To ensure compatibility with the latest reference gene annotations (ENSEMBL release 111), we removed any ncORFs that were newly classified as annotated CDSs. A manual inspection of RPF signals using the GWIPS-viz Ribo-Seq track68 in the UCSC Genome Browser69 revealed additional spurious ncORFs overlapping in-frame with CDSs from other transcripts. These were likely misidentified due to shared translation signals and were conservatively excluded using custom scripts, leaving 11,623 clusters in humans and 16,485 in mice.

To facilitate downstream analyses, a representative ncORF was selected for each cluster using a two-step strategy. First, a representative transcript was selected by prioritizing those with matched annotation from NCBI and EMBL-EBI (MANE)43 (for humans) or APPRIS principal70; otherwise, the longest transcript was chosen. Then, among ncORFs within the representative transcript, a single ncORF was selected based on start codon hierarchy (AUG > CUG/GUG > UUG) following empirical evidence28,37. If multiple ncORFs shared the same preferred start codon, the longer one was chosen.

For validation, our ncORFs were compared with previously annotated GENCODE ncORFs2 and those with MS evidence compiled previously20. Tier annotations of GENCODE ncORFs were obtained from a recent study29. Known functional ncORFs and their genomic coordinates were manually curated from published literature (Table S4). To account for possible proteoforms arising from alternative initiation and splicing, we considered an ncORF to be recovered if either its start or stop codon was identical to any ncORF in our annotation.

ncORF sequence analyses

The secondary structure surrounding each ORF (ncORF or CDS) start codon was predicted with RNAfold71 on a 150 nt window extracted from the transcript in which the ncORF resides starting from the 30th nucleotide upstream of the start codon. For each position within this window, the overall pairing probability was calculated by dividing the number of ORFs with a base-paired residue at that position by the total number of ORFs analyzed. The tRNA adaptation index and effective number of codons for ncORFs and CDSs were calculated using the “cubar” R package72. tRNA gene annotations in humans and mice were obtained from GtRNAdb (release 22)73.

Putative ncEPs analyses

Known domains within ncEPs were identified using InterProScan (v5.64-96.0)74 with default parameters. Repetitive sequence annotations were downloaded from the UCSC Table Browser. BEDTools (v2.31.1)75 was used to identify TE-overlapping ncORFs. IDRs were predicted with IUPred376 with analysis type set to “short disorders” and smoothing disabled. For each ORF, residues with disorder tendency score > 0.5 were considered disordered, and the IDR fraction was calculated as the proportion of disordered residues relative to the ORF length.

Gnocchi Score and PhyloP of ncORFs and CDSs

Genome-wide Gnocchi scores computed using 1 kb sliding windows (with 100bp step size) were downloaded from the gnomAD database50. For each base pair, the Gnocchi score was defined as the average across the ten overlapping 1kb windows that covered that position using BEDTools. The final Gnocchi score for each ORF (ncORF or CDS) was obtained by averaging the scores of all base pairs within the ORF.

Base-wise PhyloP scores computed across 441 mammals or 239 primates51 were downloaded from the UCSC Genome Browser. For each ncORF or CDS, PhyloP values were extracted for a ±30 bp window surrounding the start and stop codons using custom scripts. These values were then averaged over different ORFs to generate metagene profiles for visualization. To evaluate three-nucleotide periodicity, the metagene profiles were subjected to autocorrelation analysis using the “ac.test” function from the “testcorr” R package77. Statistical significance was assessed using the robust t̃ statistics.

ncORF origination node and mode

Whole genome alignments and corresponding species trees were downloaded from the UCSC Genome Browser (GRCh38 100-way for humans and GRCm39 35-way for mice). The origination node and mode for each ncORF were inferred following previous approaches34,35. For each ORF, the sequence from the reference genome and its orthologous regions in other species were extracted. Orthologous sequences covering less than 50% of the ncORF length relative to the reference species were excluded. The remaining sequences were realigned with MAFFT (v7.520)78 with default parameters. The corresponding gene tree was derived by pruning the species tree using Gotree (v0.4.3)79 with default settings. Ancestral sequence reconstruction was performed using FastML (v3.11)80 with parameters “-- seqType nuc --SubMatrix HKY --OptimizeBL no --indelReconstruction ML”. An ORF structure was considered intact in an orthologous or ancestral sequence if one of the first three codons was a start codon and at least 70% of the 5’ region of the reference ncORF was free of in-frame stop codons. The origination node was defined as the most ancient ancestral node that retained an intact ORF structure. An ncORF was classified as de novo if the origination node had an ancestor covering at least 50% of the ncORF sequence but lacking an intact ncORF structure. Otherwise, the origination mode was classified as uncertain. The distance from an ancestral node to the reference species (tip node) was calculated by summing branch lengths along the connecting path using the ape R package (v5.8)81.

ncORF BLS and PhyloCSF

For each ncORF, global BLS was calculated as the ratio of the total branch length of the subtree connecting all leaves with intact ORF structures to the total branch length of the full species tree as in our previous study48 with custom scripts. Local BLS was calculated similarly, but the denominator was replaced by the total branch length of the subtree defined by the origination node and all of its descendant nodes (including leaves). To evaluate coding potential, the alignment of all species descending from the origination node was converted into Multiple Alignment Format after removing the stop codon. PhyloCSF scores were then computed using the “score-msa” module of PhyloCSF++82.

ncORF expression analysis

To quantify translation levels of ncORFs and CDSs in Ribo-Seq libraries, we assigned uniquely mapped reads to corresponding P-site positions with the PSite package. The read count for each ORF was calculated as the total number of RPFs whose P-sites fell within the ORF. To reduce bias caused by RPF enrichment at ORF termini, reads mapping to start or stop codons were excluded. For CDSs, only the count of the representative transcript of each gene as defined above was used for downstream analyses. For samples with multiple libraries, read counts were aggregated prior to normalization. Sample-specific size factors were estimated using CDS counts and applied to normalize both ncORF and CDS counts across samples using DESeq283. RPKM for an ORF was calculated as normalized read counts / ORF length (nt) / library size × 109, where library size was the median of total read counts across samples to ensure cross-sample comparability. An ORF (ncORF or CDS) was considered translated if its RPKM was ≥ 1. To evaluate tissue specificity of ncORF translation, the normalized RPF count Xk of an ORF in sample k was log-transformed as Xk = log(Xk + 1). Then, tissue specificity index84 was calculated as , where and n is the total number of samples. To quantify RNA levels, raw reads of deep transcriptome profiling across different human or mouse tissues from previous studies55,56 were downloaded from SRA. Gene expression levels (transcript per million, TPM) were estimated using salmon (v1.10.3)85 with parameters “--seqBias --gcBias –posBias-l A”.

WGCNA analysis58 of ncORF and CDS translation was performed using the PyWGCNA package (v2.0.5)86 with default settings. Only ORFs with normalized RPF counts ≥ 10 in more than half of the samples were included. The soft threshold was determined automatically by PyWGCNA. Due to differences in soft thresholds, two ORFs were considered cotranslated if their topological overlap > 0.1 (human) or > 0.01 (mouse).58 To test whether co-translation is enriched between ncORFs and CDSs from the same transcript, gene identities of CDSs were randomly permuted among all co-translated ncORF–CDS pairs. This procedure was repeated 1,000 times, and the empirical P value was calculated as (m + 1) / 1,000, where m is the number of permutations in which the number of same-gene pairs was greater than or equal to the observed count. The bipartite network of ncORF-CDS co-translation was clustered using the Leiden algorithm87 implemented in igraph (v2.1.3)88 by optimizing modularity with a resolution of 0.2. The top 20% of nodes by betweenness centrality were designated as hub nodes. Gene ontology analysis of hub genes in the largest modules was conducted using ClusterProfiler (v4.12)89.

eLife Assessment

This study presents a large, systematically curated catalog of non-canonical open reading frames (ncORFs) in human and mouse by reanalyzing nearly 400 Ribo-seq datasets using a standardized pipeline; the resulting atlas consolidates ncORF annotations across tissues and provides a valuable reference for understanding non-canonical translation and ORF emergence. The main conclusions are supported by consistent data processing and multiple computational measures of translation and conservation. While the pipeline is transparent and robust, several downstream analyses are descriptive, and some evolutionary interpretations remain correlative; dataset heterogeneity, uneven tissue representation, and limited experimental validation also constrain the strength of a subset of the findings. Overall, the evidence is solid, and the resource will be broadly used by the community.

Reviewer #1 (Public review):

This work compiles a comprehensive atlas of ncORFs across mammalian tissues and cell types, derived from reanalysis of ~400 public ribosome profiling datasets. The authors then evaluate cross-species conservation and functional signatures, proposing that evolutionarily ancient ncORFs tend to have higher translation potential, stronger expression, and closer relationships with canonical coding sequences.

Strengths:

In general, the study provides a large-scale and timely resource of annotated ncORFs, which could be broadly useful for the community. The authors collected ~400 public ribosome profiling datasets for annotations of ncORFs, which, to my best knowledge, is the largest collection of data for such a purpose. The catalog could facilitate future investigations into ncORF biology and broaden understanding of the coding potential of the "non-coding" genome.

Weaknesses:

Based on the ncORF catalog, some of the analyses were not properly done. Some of the results are descriptive.

(1) Bias and representations of the data source. Public ribo-seq datasets are unevenly distributed across tissues and cell lines, raising concerns about heterogeneity and underrepresentation of certain contexts. This may limit the generalizability of the catalog.

(2) The discussion on modular domains of ncORFs is unclear, and the claim that they may originate via TE-related mechanisms is not well supported. Stronger evidence or clearer reasoning is needed.

(3) The conservation comparisons are not fully convincing. Figure S7 shows only mild differences between ncORFs and CDS, and statistical significance is not clearly demonstrated.

Comparisons with other non-coding RNAs should be added, and overlapping sequences between ncORFs and CDS should be excluded to avoid bias.

(4) Figure 3 indicates that some ncORFs are subject to evolutionary constraints. This is not surprising. The authors should provide further analyses on more detailed features of these "conserved" ncORFs vs. the "non-conserved" ones. Some pretty informative works have been done in Drosophila, worms, mice, and humans. Figure 3 suggests some ncORFs are under evolutionary constraint, but this is not unexpected. More granular analyses contrasting "conserved" versus "non-conserved" ncORFs would be informative. In fact, small ORFs, especially uORFs, have been extensively studied for their functions and cross-species conservation. The authors should explicitly show what is new here in their analyses.

(5) Translation levels are reported using RPF counts. However, translation efficiency (normalized by RNA expression) is a more appropriate measure to account for expression heterogeneity.

(6) The correlation analyses between ncORF translation levels and PhyloCSF are confusing and largely descriptive. These sections need sharper framing and clearer conclusions.

(7) Public ribo-seq datasets, generated by different research labs, are known for their strong batch effects. Representations of tissues and cells are also very unbalanced. Therefore, the co-translation analysis between ncORFs and canonical CDS is not well controlled. This should be done by referring to a recent large-scale ribo-seq meta-analysis (Nat Biotechnol. 2025. doi: 10.1038/s41587-025-02718-5).

Reviewer #2 (Public review):

Summary:

Chang et al. attempted to analyze a large number of ribo-seq datasets through a standardized pipeline, identifying novel non-canonical ORFs and elucidating their evolutionary and expression characteristics.

Strengths:

(1) The datasets analyzed by the authors are sufficiently comprehensive, and the use of standardized pipelines ensures excellent analytical consistency.

(2) Their analyses of ORF evolution and co-expression further deepen our understanding of these ORFs.

Weaknesses:

(1) The authors primarily conducted analyses through bioinformatics, lacking sufficient wet-lab experimental evidence.

(2) Regarding the evolution of non-canonical ORFs, a considerable amount of prior work already exists. The authors need to further clarify what new insights and discoveries they have made based on the analysis of such a large dataset.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

This work compiles a comprehensive atlas of ncORFs across mammalian tissues and cell types, derived from reanalysis of ~400 public ribosome profiling datasets. The authors then evaluate cross-species conservation and functional signatures, proposing that evolutionarily ancient ncORFs tend to have higher translation potential, stronger expression, and closer relationships with canonical coding sequences.

Strengths:

In general, the study provides a large-scale and timely resource of annotated ncORFs, which could be broadly useful for the community. The authors collected ~400 public ribosome profiling datasets for annotations of ncORFs, which, to my best knowledge, is the largest collection of data for such a purpose. The catalog could facilitate future investigations into ncORF biology and broaden understanding of the coding potential of the "non-coding" genome.

We thank the reviewer for the positive evaluation of our manuscript and for recognizing the significance of our contribution.

Weaknesses:

Based on the ncORF catalog, some of the analyses were not properly done. Some of the results are descriptive.

(1) Bias and representations of the data source. Public ribo-seq datasets are unevenly distributed across tissues and cell lines, raising concerns about heterogeneity and underrepresentation of certain contexts. This may limit the generalizability of the catalog.

We agree with the reviewer that the uneven distribution of public Ribo-seq datasets across tissues can inevitably introduce bias in the ncORF composition of our catalog. This bias is likely more pronounced in humans due to the narrower tissue coverage. We have addressed this point in the Discussion section of the revised manuscript.

(2) The discussion on modular domains of ncORFs is unclear, and the claim that they may originate via TErelated mechanisms is not well supported. Stronger evidence or clearer reasoning is needed.

We thank the reviewer for highlighting this point. We have revised the manuscript to more clearly explain the rationale behind our analysis of ncORF modular domains and have adopted more cautious language regarding their potential transposable element–related origins, limiting interpretations to what is directly supported by the data.

(3) The conservation comparisons are not fully convincing. Figure S7 shows only mild differences between ncORFs and CDS, and statistical significance is not clearly demonstrated.

Comparisons with other non-coding RNAs should be added, and overlapping sequences between ncORFs and CDS should be excluded to avoid bias.

We thank the reviewer for this comment and apologize for the lack of clarity in the original figure. Both CDSs and ncORFs show significant deviation from zero Gnocchi scores (two-sided Wilcoxon signed-rank tests), which is now stated explicitly in the revised legend and text. CDS-overlapping ncORFs were already excluded in the original analysis; this has been clarified to avoid confusion.

As suggested, we have added lncRNAs for comparison. ncORFs display modestly higher Gnocchi scores than lncRNAs, and this difference persists when restricting the analysis to lncRNA-derived ncORFs and their corresponding full-length lncRNAs (see revised Fig. S7). These additions strengthen the conservation comparison while controlling for transcript context.

(4) Figure 3 indicates that some ncORFs are subject to evolutionary constraints. This is not surprising. The authors should provide further analyses on more detailed features of these "conserved" ncORFs vs. the "non-conserved" ones. Some pretty informative works have been done in Drosophila, worms, mice, and humans. Figure 3 suggests some ncORFs are under evolutionary constraint, but this is not unexpected. More granular analyses contrasting "conserved" versus "non-conserved" ncORFs would be informative. In fact, small ORFs, especially uORFs, have been extensively studied for their functions and cross-species conservation. The authors should explicitly show what is new here in their analyses.

We thank the reviewer for this insightful comment. We agree that cross-species conservation of ncORFs (particularly uORFs) has been extensively investigated in prior studies, including our own.

However, most prior analyses have focused on conservation of start codons or overall ORF integrity, which does not distinguish selection acting on translational activity from selection acting on the encoded peptide sequence itself. In contrast, our analysis leverages codon-level periodic PhyloP signals across the full ORF. The observed three-nucleotide periodicity is consistent with selective constraint at the amino acid level, rather than merely preservation of initiation sites or translational potential. Furthermore, our newly developed branch-length statistic uncovers lineage-restricted conservation patterns among ncORFs, enabling resolution of evolutionary dynamics not captured by conventional conservation metrics.

Thus, while the existence of conserved ncORFs is not unexpected, the conceptual advance of our study lies in demonstrating that a subset exhibits coding-like evolutionary constraint consistent with selection on their peptide products, as well as revealing lineage-specific conservation patterns. We have clarified this distinction in the revised Discussion.

(5) Translation levels are reported using RPF counts. However, translation efficiency (normalized by RNA expression) is a more appropriate measure to account for expression heterogeneity.

We agree that translation efficiency (TE), which normalizes ribosome footprint counts by RNA abundance, is in principle an appropriate metric. We initially calculated TE and compared ncORFs with CDSs. However, we found that TE estimates for short ncORFs were substantially inflated by RPF enrichment near start and stop codons, leading to unstable and potentially misleading values.

For CDSs, this bias is commonly addressed by excluding the first and last 10 to 20 codons when quantifying RPF density. This strategy is not feasible for ncORFs because of their short length. We therefore used RPF counts in the final analysis, applying stringent positional filtering. Only RPFs whose P sites fall within the ORF body, excluding start and stop codons, were counted. RPFs overlapping the ORF but with P sites outside the annotated frame, likely derived from adjacent ORFs or initiation or termination pausing, were excluded.

TE and RPF counts both measure translation but capture different aspects. TE reflects ribosome density relative to transcript abundance, whereas RPF counts quantify overall ribosome engagement. Given the short lengths of ncORFs, count-based quantification provides a more robust and conservative estimate of their translational activity.

(6) The correlation analyses between ncORF translation levels and PhyloCSF are confusing and largely descriptive. These sections need sharper framing and clearer conclusions.

We thank the reviewer for this comment. We agree that the original presentation lacked clear framing. The relationship between PhyloCSF scores and mean ncORF translation levels across tissues is influenced by both evolutionary age and tissue specificity. Older ncORFs with higher coding potential tend to exhibit stronger tissue-restricted expression. As a result, their mean translation levels across all tissues appear lower, not because they are weakly translated, but because their translation is concentrated in specific tissues. This point is addressed in the revised manuscript.

(7) Public ribo-seq datasets, generated by different research labs, are known for their strong batch effects. Representations of tissues and cells are also very unbalanced. Therefore, the co-translation analysis between ncORFs and canonical CDS is not well controlled. This should be done by referring to a recent large-scale ribo-seq meta-analysis (Nat Biotechnol. 2025. doi: 10.1038/s41587-025-02718-5).

We thank the reviewer for highlighting this important study and for raising concerns regarding batch effects and tissue imbalance in public Ribo-seq datasets. We are aware that public Ribo-seq data generated by different laboratories are subject to substantial batch effects. During the ncORF annotation phase, we applied stringent quality-control criteria to minimize technical variability. For the co-translation analysis, inclusion criteria were relaxed to increase tissue and cell-type coverage. To partially mitigate representation bias, libraries derived from the same tissue or cell type were merged when quantifying ORF translation levels, thereby reducing overrepresentation from heavily sampled contexts.

Nevertheless, we acknowledge that these measures cannot completely eliminate batch effects or imbalance inherent to public datasets. We agree that co-translation analysis would benefit from uniformly processed, high-quality datasets generated under standardized protocols with balanced tissue representation, representing a valuable direction for future research.

Reviewer #2 (Public review):

Summary:

Chang et al. attempted to analyze a large number of ribo-seq datasets through a standardized pipeline, identifying novel non-canonical ORFs and elucidating their evolutionary and expression characteristics.

Strengths:

(1) The datasets analyzed by the authors are sufficiently comprehensive, and the use of standardized pipelines ensures excellent analytical consistency.

(2) Their analyses of ORF evolution and co-expression further deepen our understanding of these ORFs.

We thank the reviewer for the positive evaluation of our manuscript. It is encouraging to know that the analytical framework was found to be sound and appropriate.

Weaknesses:

(1) The authors primarily conducted analyses through bioinformatics, lacking sufficient wet-lab experimental evidence.

We thank the reviewer for this comment and acknowledge this limitation. We agree that functional validation through wet-lab experiments would provide important mechanistic insight into individual ncORFs. However, this study was designed as a systematic, genome-wide computational analysis to characterize translated ncORFs across species and tissues. Our objective was to define global patterns of translation, conservation, and structural features using large-scale datasets. Given the breadth and scale of these analyses, experimental validation of specific ncORFs falls beyond the scope of the current study. We have clarified this point in the dicussion and noted that our results provide a framework for future targeted experimental investigation.

(2) Regarding the evolution of non-canonical ORFs, a considerable amount of prior work already exists. The authors need to further clarify what new insights and discoveries they have made based on the analysis of such a large dataset.

We thank the reviewer for this suggestion. Similar concerns were also raised by Reviewer #1. In response, we have revised the Discussion to more clearly delineate the conceptual advances enabled by our large-scale dataset.

Recommendations for the authors:

Reviewing Editor Comments:

Several aspects of the downstream analyses would benefit from additional refinement. The heterogeneity and tissue imbalance inherent in public Ribo-seq datasets introduce potential biases in ncORF detection and inferences about co-translation. Given the breadth of the dataset, it would also be informative to quantify how consistently the newly identified ncORFs are detected across samples-distinguishing those observed broadly across tissues, those enriched in specific contexts, and those detected in only a few datasets. Such stratification would help differentiate reproducibly translated ORFs from candidates requiring further validation.

We thank the editor for the helpful comments. We agree that heterogeneity and tissue imbalance in public Ribo-seq datasets can influence ncORF detection and downstream interpretations. We have added discussion of this limitation in the revised manuscript.

Detection of ncORF translation depends not only on biological activity but also on sequencing depth and data quality. Although all ncORFs reported here were reproducibly identified by multiple methods across independent libraries, we agree that those detected in a larger number of datasets represent stronger candidates for functional validation. Accordingly, we now report the number of methods and libraries in which each ncORF was detected in the final catalog (Supplementary Table 3). Overall, 22.3–26.3% of ncORFs were detected in more than 10 libraries, whereas more than half were observed in only two to five libraries (Fig. S1B), enabling clearer stratification of broadly translated versus more context-specific candidates.

Some evolutionary and functional interpretations are largely descriptive or consistent with established findings for small ORFs, and the authors should more clearly articulate what is novel in their analyses. The criteria separating "young," "old," and "ancient" ORFs require clearer definition, and conservation analyses would be strengthened by improved statistical rigor and explicit exclusion of regions overlapping annotated coding sequences. Evidence for modular domain features or transposable element-related origins is limited and warrants either stronger support or more cautious framing. Proteomics validation is currently minimal and could be substantially reinforced using existing public MS resources.

We thank the reviewer for these constructive comments. In the revised manuscript, we more clearly delineate the novel insights derived from our evolutionary analyses of ncORFs, distinguishing them from established findings on small ORFs.

We have clarified the criteria used to classify ORFs by evolutionary age in figure 6E and refined the terminology describing “young,” “old,” and “ancient” categories to ensure precise definition. The conservation analyses have been strengthened through more rigorous statistical treatment and by explicitly excluding regions overlapping annotated coding sequences.

With respect to modular domain features and potential transposable element–related origins, we have adopted more cautious language and limited our interpretations to what is directly supported by the data. Finally, we acknowledge that current proteomic validation remains limited and have clarified this point in the manuscript while outlining the potential for future integration of large-scale public mass spectrometry datasets in Discussion.

The authors additionally report an interesting observation that many ncORFs on mRNA co-translate with the main CDS of the same gene. Because canonical models often posit that uORF translation suppresses downstream CDS translation, further analysis would be valuable. In particular, it would be useful to determine whether patterns of co-translation differ among ORF types or evolutionary categories and to discuss possible regulatory mechanisms underlying these relationships.

We thank the editor for this thoughtful comment. As noted in our response to Reviewer #2, uORF–CDS co-translation does not contradict the canonical model in which uORFs repress downstream CDS translation. Co-translation reflects concurrent ribosome occupancy, whereas repression concerns the fraction of initiating ribosomes that ultimately reach and translate the CDS. Following the editor’s suggestion, we further examined whether co-translation patterns differ across ORF types or evolutionary categories. We found that ncORFs co-translating with their corresponding main CDSs are predominantly uORFs. However, these uORFs do not show statistically significant differences in conservation metrics or evolutionary age compared with other non-overlapping uORFs. Thus, we did not detect clear subtype- or age-specific distinctions among co-translating ncORFs. We have clarified these analyses in the revised manuscript.

Addressing these points would enhance the precision, interpretability, and robustness of the study's conclusions.

Reviewer #2 (Recommendations for the authors):

(1) The authors developed and refined a standardized pipeline to analyze nearly 400 ribo-seq datasets, identifying over 10,000 novel non-canonical ORFs in both human and mouse samples. Given the scale of this analysis, it is intriguing to consider how many of the newly identified non-canonical ORFs are consistently detected across multiple sample types (conservatively expressed ORFs), how many are restricted to specific tissues/ or tissue-specific ORFs), and how many were detected in only a single or very few samples (ORFs requiring further validation). Providing these data could offer new insights into understanding ORF translation.

Thanks for this constructive suggestion. This information has been presented in the revised Supplementary Table 3 and in a newly added supplementary figure (Fig. S1B), which together provide a clearer overview of ncORF detection consistency and context specificity.

(2) The authors' validation of MS data lacks specific details in the paper. Regarding the MS-supported ORF mentioned in Lane 117, which dataset's MS data is being referenced? Or does it refer to the content in Reference 20? At present, substantial research exists in both public general proteomics studies (e.g., CPTAC) and MS investigations targeting non-canonical ORFs. We recommend the authors incorporate additional MS data or public MS-based databases to strengthen validation in this area (PMID: 34129944, 39794466, 37823596，39413795).

We thank the reviewer for this comment and for the helpful suggestions. The MS-supported ORFs mentioned in line 117 refer to the compilation reported in Reference 20, which integrates evidence from multiple independent proteomics studies. In addition, we examined MS-supported ORFs curated by GENCODE and PeptideAtlas, which are shown in Fig. 1E.

We agree that incorporating additional MS datasets would further strengthen validation of ncORFs. Studies cited by the reviewer and recent community efforts such as the GENCODE and PeptideAtlas analyses (PMID: 39314370) provide valuable examples in this direction. However, performing a comprehensive reanalysis of more than 95,000 public human MS runs is computationally demanding and currently infeasible for our group given resource and funding constraints.

To our knowledge, ongoing community-wide initiatives are working toward more comprehensive catalogs of translated human ncORFs. Large-scale, exhaustive MS searches will be particularly effective once a community consensus annotation framework for ncORFs is established. We have added discussion of these limitations and future directions in the revised manuscript.

(3) The authors classified ncORFs into three groups-"Ancient," "Young," and "Old"-based on their origin nodes. However, both the "Young" and 'Old' groups appear to be "mammalian-specific," yet the specific criteria for their division remain unclear. It is recommended to more clearly define in the figure legend or main text how "Young" and "Old" are categorized (e.g., based on specific evolutionary nodes or distance thresholds from nodes to the end) to avoid reader confusion.

In Fig. 5, “old” and “young” were intended as qualitative descriptors of relative evolutionary age based on the position of ncORF origination nodes along the phylogeny, as indicated on the x-axis. They were not meant to represent discrete categories. To avoid confusion, we have revised the manuscript to use “older” and “younger” throughout when referring to relative age differences. A binary classification is used only in Fig. 6E, where ncORFs are grouped into ancient (pre-mammalian) and younger (mammalian-specific) categories. This distinction is clearly defined in both the main text and the corresponding figure legend.

(4) The authors observed an intriguing phenomenon: ncORFs on mRNA tend to co-translate with the main CDS of the same gene. However, the conventional view holds that uORF translation often inhibits the translation of the main CDS. I suggest the authors could refine their analysis in this section further. For instance, do different types of ORFs or ORFs at different evolutionary levels exhibit distinct levels of cotranslation with the main CDS? Additionally, while observing this phenomenon, the authors should also propose hypotheses regarding the regulatory mechanisms involved in these processes.

We thank the reviewer for these constructive suggestions. After excluding CDS-overlapping ORFs, we identified 258 human and 128 mouse ncORFs that co-translate with their corresponding main CDSs. With the exception of 10 human dORFs, all remaining cases were uORFs. We compared these cotranslating ncORFs with other non-overlapping uORFs and dORFs but did not detect statistically significant differences in evolutionary age and conservation metrics. Because no clear distinguishing features emerged, we did not include these results in the manuscript.

Importantly, the observation of uORF–CDS co-translation does not contradict the established repressive role of uORFs. Co-translation reflects concurrent ribosome occupancy, whereas repression concerns the proportion of initiating ribosomes that ultimately translate the CDS. For example, if two ribosomes initiate within a given interval and one translates the uORF while one translates the CDS, CDS output is reduced by 50% relative to a uORF-free transcript. If four ribosomes initiate under the same repressive regime, two may translate the uORF and two the CDS. In this case, absolute translation of both ORFs increases, while the fractional repression remains unchanged. Thus, co-translation is compatible with a regulatory model in which uORFs reduce CDS translation efficiency without abolishing it. This has been clarified in the revised manuscript.

Version 2

Abstract

Non-canonical open reading frames (ncORFs) are pervasive within transcripts annotated as “non-coding” or “untranslated regions” of mRNAs, yet their landscape under normal physiological conditions remains to be fully resolved, particularly outside humans. Here we applied a stringent and standardized pipeline to hundreds of high-quality ribosome profiling libraries from normal mammalian tissues and cell types, identifying 11,623 human and 16,485 mouse ncORFs. Evolutionary analyses revealed that thousands of ncORFs are subject to coding constraint and exhibit lineage-specific conservation, underscoring their functional potential. Ancient ncORFs are preferentially highly translated, broadly expressed, and enriched for lineage-specific conservation. Co-expression patterns further indicate that many ncORFs, especially ancient ones, are cotranslated with canonical coding sequences, consistent with functions mediated through protein–protein interactions. Together, these findings establish a comprehensive atlas of mammalian ncORFs and provide fundamental insights into their evolutionary dynamics and functional integration within the proteome.

Introduction

The systematic identification of functional genomic elements represents a foundational objective in the field of genomics. Among these elements, non-canonical open reading frames (ncORFs)—which are open reading frames (ORFs) located within long non-coding RNAs (lncRNAs) or untranslated regions (UTRs) of mRNAs—have attracted considerable attention in recent years1-5. While certain ncORF subtypes, such as upstream ORFs (uORFs) located in 5’ UTRs and downstream ORFs (dORFs) in 3’ UTRs, have been extensively characterized for their ability to regulate mRNA translation and stability in cis, a growing body of evidence demonstrates that ncORFs can also give rise to functional proteins that engage in diverse cellular pathways1,2,6,7. As functional ncORFs continue to be discovered across multiple eukaryotes8-17, systematic elucidation of their composition and function promises to uncover an additional layer of regulatory complexity and to refine our conception of the so-called “non-coding” genome.

Although ncORFs may play important biological roles, the number of translated ncORFs in the human genome remains uncertain. Detection by mass spectrometry (MS)-based proteomics is challenging, likely due to their short lengths and low abundances18. Ribosome profiling (Ribo-Seq) offers greater sensitivity for identifying translation events18,19, but downstream computational pipelines often produce highly divergent ncORF sets 20. Consequently, published estimates of ncORF abundance span several orders of magnitude—ranging from thousands to millions 15-22. To establish a standardized reference, the GENCODE consortium integrated data from seven distinct studies to produce a catalog of 7,264 human ncORFs (hereafter GENCODE ncORFs)2; however, the heterogeneity of sample sources and prediction algorithms likely introduced systematic bias. This catalog is also restricted to AUG-initiating ORFs, thereby overlooking those with near-cognate start codons. A later effort generated a high-resolution human translatome map and identified 7,767 ncORFs with a unified pipeline22. Nevertheless, this conservative annotation disproportionately under-represents coding sequence (CDS)-overlapping ncORFs. Indeed, later estimation suggests that between 10,500 to 22,500 ncORFs should exist in the human genome18. Furthermore, many ncORFs were reported from studies of cancer cells1,13,23,24, where extensive transcriptome reprogramming alters the translational landscape25,26. While paired healthy tissues were also analyzed in these studies, the landscape of ncORFs translated under normal physiological conditions is not yet fully resolved.

While a limited subset of ncORFs has been shown to encode signaling molecules, cofactors, membrane proteins, or subunits of large protein complexes1,27,28, functions of most ncORF-encoded proteins (ncEPs) are largely uncharacterized. Advances in high-throughput screen technologies have enabled the systematic interrogation of functional ncEPs23,28-32. Nevertheless, there is minimal concordance among hits from different studies3. Evolutionary analyses can offer insights into potential functionality of these translation products and mechanisms of origination33-35, yet how ncORF translation has been remodeled to integrate into proteome since their evolutionary origins is still poorly understood. Consequently, the extent to which translated ncORFs encode functional proteins and the evolutionary dynamics of their integration into the existing proteome are largely unresolved. Moreover, because most ncORF studies have focused on humans, it is unclear whether ncORFs in other species share similar sequence features and expression patterns.

To address these gaps, we generated a standardized annotation of translated ncORFs by uniformly reprocessing ∼400 high-quality ribosome profiling libraries from normal mammalian tissues and cell types. This stringent approach yielded 11,623 human and 16,485 mouse ncORFs, consistently recovered across diverse samples and prediction strategies. Comparative and functional genomic analyses revealed that thousands of these ncORFs are under evolutionary constraint and exhibit distinct translation profiles. The resulting dataset provides a robust, comprehensive resource that will facilitate future investigations into the biological roles of ncORFs across cellular and organismal contexts.

Results

Annotation of ncORFs translated in normal mammalian cells

We developed a standardized pipeline to systematically annotate ncORFs translated in normal human and mouse cells (Fig. 1A). We first curated publicly available Ribo-Seq libraries, excluding those from pathological or genetically modified samples. Since both library size and the proportion of in-frame ribosome-protected fragments (RPFs) affect the sensitivity and accuracy of ncORF prediction20, we filtered out libraries containing fewer than 5 million RPFs or with in-frame rates below 60%. After this quality control step, we retained 174 human and 209 mouse high-quality Ribo-Seq libraries, spanning a range of tissues and cell types (Fig. S1A; Table S1). We predicted translated ORFs that begin with NUG (where N stands for any nucleotide) start codons and encode at least five amino acids using three tools—PRICE36, RiboCode37 and Ribo-TISH38—which we previously benchmarked for optimal performance20. After excluding annotated CDSs and their extension or truncation variants arising from alternative initiation sites, we obtained a preliminary list of ncORFs in each library.

Annotation of ncORFs in humans and mice.

(A) Overview of the integrative pipeline for ncORF annotation in human and mouse genomes. (B-C) RPF coverage plot for a representative uORF (B) and a lncORF (C), visualized using the Ribo-Seq signal track from GWIPS-viz72. ncORFs were shown in yellow and CDSs in cyan. Dashed lines indicate identical genomic regions shared across transcript isoforms. (D-E) Proportion of GENCODE ncORFs rediscovered in this study. GENCODE ncORFs were stratified either by reproducibility across independent studies (D) or by tiered translation evidence strength (E). (F) Known ncORFs that were experimentally characterized in earlier studies and independently rediscovered in this study from humans and mice. CDS, coding sequence; FLOSS, fragment length organization similarity score; MANE, matched annotation from NCBI and EBI; ncORF, non-canonical open reading frame; ORF, open reading frame; RPF, ribosome-protected fragment.

To generate reliable ncORF annotations, we applied stringent filtering and integration criteria. We removed potential false positives caused by non-ribosomal contamination or translational noise based on metrics including Fragment Length Organization Similarity Score (FLOSS)39, Ribosomal Release Score (RRS)40,41 and the number of RPFs within an ORF by P-site. We then aggregated the remaining ORFs to identify those consistently detected across multiple methods and libraries in human and mouse datasets, respectively (Fig. 1A). In many cases, two ORFs predicted from different libraries or methods may share the same start or stop codon due to the alternative translation initiation or splicing, resulting in groups of overlapping ncORFs. For example, 11.1% (805) of GENCODE ncORFs overlap with at least one other ORF, forming 329 overlapping groups (Table S2). This overlap complicates the identification of reproducible ORFs, as ORFs within a group may represent distinct proteoforms of a protein. To address this issue, we collapsed each group of ncORFs sharing the same start or stop codon into a single cluster, and retained only those clusters supported by at least two independent methods and observed in more than one library. For each retained cluster, we selected a representative ORF by prioritizing transcript annotation42,43, start codon type, and ORF length to facilitate downstream analyses (Fig. 1A).

This approach yielded 12,469 translated ncORFs in humans and 16,960 in mice. However, manual inspection revealed a subset of spurious ncORFs that are not truncation or extension variants of annotated CDSs, but instead partially overlap with annotated CDSs from alternative transcript isoforms in frame (Figs. 1B & C). For instance, in gene AGTPBP1, the CDS of transcript ENST00000357081 begins downstream of ENST00000628899 CDS due to the inclusion of an alternative exon. As a result, the 5’ UTR of ENST00000357081 appears to contain a translated uORF (Fig. 1B), which overlaps in-frame with the CDS of ENST00000628899 and may be mistakenly identified as a translated ncORF owing to active translation of the latter. A similar case is seen with the lncRNA ORF (lncORF) in ENST00000462211, which overlaps in-frame with the CDS of ENST00000328767 from the TMTC4 gene and may thus be erroneously identified as a translated ncORF (Fig. 1C). To avoid such artifacts, we implemented an additional filtering step to exclude these overlapping cases and left with 11,623 and 16,485 high-confidence ncORFs in humans and mice, respectively (Table S3). Notably, while more than half ncORFs were detected in two to five libraries, 22.3–26.3% were identified in more than 10 libraries, indicating a substantial subset with robust and recurrent translation signals suitable for future experimental validation (Fig. S1B).

Annotated ncORFs are supported by independent evidence

To assess whether the annotated ncORFs are supported by independent evidence, we compared them with a previously compiled catalog of ncORFs identified using mass spectrometry (MS)-based proteomic data20. For the 3,494 MS-supported ncORFs in humans and 873 in mice, 44.5% and 67.6%, respectively, were also predicted as translated ncORFs in this study (Fig. S2). We further evaluated our annotation against the GENCODE ncORFs2, a widely used community resource. Of the 7,264 GENCODE ncORFs, 6,974 are located in fully annotated transcripts that were included in our analysis. Among them, 2,997 were reported by at least two independent studies (hereafter “phase 1”), while the remaining were only reported in a single study (“single”)2. We found that 75.6% of phase 1 ncORFs were independently recovered in our annotation, significantly more than the single-study ncORFs (Fig. 1D; P = 6.4×10-307, Fisher’s exact test).

A recent study conducted a comprehensive pan-proteome analysis to provide high-confidence peptide evidence for GENCODE ncORFs44. In this study, ncORFs were categorized into four tiers: tier 1 and 2 ncORFs had support from both MS and Ribo-Seq data; tier 3 had MS evidence only; and tier 4 had Ribo-Seq evidence only45. Notably, our annotation was significantly enriched for tier 1 or 2 ncORFs (P = 2.0×10-64, Fisher’s exact test), with 446 and 661 of them detected, respectively (Fig. 1E). Combining MS-supported ncORFs from our previous compilation and those from GENCODE ncORFs, 2,088 (18.0%) human ncORFs identified in this study were supported by MS evidence. Despite this overlap, the majority (65.2%) of human ncORFs in our annotation are not included in either of the two catalogs. This highlights both the complexity of the non-canonical translatome and the critical role of tissue diversity in comprehensive ncORF annotation.

To further validate our findings, we manually curated ncORFs with known functions from the literature. We identified 17 ncORFs in our annotation that have been experimentally confirmed to be translated—either through reporter assays or antibody detection—and shown to produce functional proteins (Fig. 1F; Table S4). Notably, 16 of these validated ncORFs are from humans, and only 12 are included in the GENCODE catalog. This analysis further underscores the accuracy and functional relevance of our annotations. Together, these results indicate that our annotation pipeline generates a robust and reliable catalog of translated ncORFs in mammals.

ncORFs exhibit signatures of translation and encode putative proteins lacking known domains

We classified ORFs into categories based on community recommendations2 and found that the distribution of ORF types closely aligns with previous estimates18. In both human and mouse, uORFs, upstream overlapping ORFs (uoORFs), and lncORFs are more prevalent than other types, whereas internal ORFs (iORFs), downstream overlapping ORFs (doORFs), and dORFs are relatively rare (Fig. 2A). Although most ncORFs are short, encoding fewer than 100 amino acids (Fig. 2B), their sequences show clear signatures of translation similar to CDSs. These include preferential usage of AUG start codons (Fig. S3A), enrichment toward 5’ transcript ends (Fig. S3B), and reduced RNA secondary structure46 (Fig. S3C) and optimized Kozak contexts47,48 around start codons, especially in non-AUG ncORFs (Figs. S3D & S4). Despite these similarities, ncORFs differ markedly from CDSs in both codon and amino acid usage (Figs. S5A-B). This divergence is further supported by their lower tRNA adaptation index values and higher effective number of codons, indicating less optimized codon usage (Figs. S5C-D). Together, these findings show that ncORFs share core sequence and structural features with canonical CDSs, yet maintain distinct codon and amino acid usage profiles.

Sequence features of ncORFs and putative ncEPs in humans and mice.

(A) Number of ncORFs belonging to different categories. (B) Distribution of ncORF lengths. (C) Proportion of ncEPs that contain known Pfam domains. (D) Number of ncORFs with Pfam domains among lncORFs and other ncORF categories. (E) Distribution of the proportion of intrinsically disordered residues in CDS- and ncORF-encoded proteins. (F) Most frequently identified Pfam domains among human and mouse ncORFs. (G) Proportion of ncORFs overlapping with TEs, stratified by presence or absence of Pfam domains. Differences were assessed using Wilcoxon rank-sum tests. CDS, coding sequence; ncORF, non-canonical open reading frame; ncEP, ncORF-encoded protein; TE, transposable element.

Although ncORFs can encode putative proteins, the functional significance of ncEPs remains poorly understood. As an initial step toward functional inference, we systematically searched for annotated protein domains within ncEP sequences, reasoning that the presence of conserved domains would suggest potential biochemical activity. Overall, only 1.29% (150) of human and 0.78% (129) of mouse ncEPs contained at least one known domain (Fig. 2C; Table S5), presumably due to their short lengths relative to canonical proteins. Consistently, domain-containing ncORFs are significantly longer than those without annotated domains (Fig. S6) and are predominantly classified as lncORFs (Fig. 2D). Unlike canonical proteins, which have relatively low intrinsically disordered region (IDR) content (median ± standard error: 14.6±0.28% in humans and 12.7±0.26% in mice), ncEPs exhibit significantly higher IDR proportions (Fig. 2E), with medians of 62.5±1.42% and 60.0±1.27%, respectively. These results indicate that most ncEPs lack structured domains and are instead dominated by IDRs.

Among the eight domains observed in at least five ncORFs, three are DNA-binding domains: zinc finger, Krüppel-associated box, and helix-turn-helix (Fig. 2F). Because transposable elements (TEs) commonly encode DNA-binding proteins required for transposition49, we tested whether domain-containing ncORFs preferentially overlap TE-derived sequences. Indeed, domain-containing ncORFs were significantly enriched for TE overlap in both humans and mice (P = 5.8×10-4 and 3.4×10-17, respectively; Fisher’s exact tests; Fig. 2G). Although enrichment alone does not establish origin, it supports a model in which a subset of longer ncORFs may have acquired modular domains through TE-derived sequence contributions.

Putative ncEPs are selectively constrained

Given the scarcity of known domains in putative ncEPs, a key question is whether they are functional in cells. If at least some of these ncEPs are not merely translational noise or byproducts of translation-dependent functions, we would expect them to exhibit evolutionary signatures of selective constraints similar to canonical CDSs. To test this, we leveraged the recent high-resolution genomic constraint maps derived from both intra- and inter-species comparisons in humans 50,51. First, we examined the genomic non-coding constraint of haploinsufficient variation (Gnocchi), which quantifies selective pressure by comparing observed versus expected variants across human genomes. A Gnocchi score above zero indicates purifying selection50. Although ncORFs exhibit lower Gnocchi scores than CDSs, their score remain significantly above zero and exceed those of lncRNAs (Fig. S7), suggesting that ncORF sequences are under evolutionary constraint. A hallmark of functional coding regions is codon position-specific constraints, but Gnocchi scores lack the resolution needed to assess such fine-scale features. Therefore, we next analyzed the base-resolution conservation metric PhyloP52 calculated from multiple alignments of primate or mammalian genomes51. As expected, PhyloP scores in CDSs are substantially higher than those in flanking UTRs and display strong three-nucleotide periodicity, with the third codon position being less conserved than first and second (Fig. 3A). Although ncORFs (excluding those overlapping annotated CDSs) do not show higher overall PhyloP than flanking regions, they also exhibit significant three-nucleotide periodicity in both mammals (Fig. 3A) and primates (Fig. S8), suggesting that at least a subset of ncORFs are subject to similar evolutionary constraints as CDSs.

Evolutionary constraints of ncORFs.

(A) The mean PhyloP scores of 30 base pairs upstream and downstream of the ORF start or stop codons in mammals. Codons were delineated with bars of alternating colors by different frames, and nucleotides in untranslated regions were shown in grey. The significance of three-nucleotide periodicity was assessed by autocorrelation with a lag of three. (B) Cladograms of vertebrates with the number of ncORFs originating at each ancestral branch of humans. Triangles indicate species merged into larger clade for visual simplicity. (C) Relationship between ORF origination rates and node ages measured as node-to-tip distances. Origination rate was defined as the number of ncORFs that originated on a branch divided by the branch length. Blue lines show linear regression fits, and grey bands represent 95% prediction confidence intervals. Spearman’s correlation is indicated. (D) Distribution of ncORF PhyloCSF scores normalized by the number of codons in per ncORF. The dashed line denotes zero. CDS, coding sequence; ncORF, non-canonical open reading frame; ORF, open reading frame.

Motivated by the above results, we next sought to identify individual ncORFs under selective pressure. To this end, we determined their modes of origination33,34 and evolutionary history through comparative genomic analyses. Our analysis revealed that most ncORFs are evolutionarily young, with 63.0% (7,319) and 82.8% (13,508) ncORFs in humans and mice originating within the mammalian lineage, respectively (Figs. 3B & S9). In line with previous studies33,34, we estimate that at least 51.7% of human and 50.8% of mouse ncORFs arise de novo (Table S6). We also observed that the number of ncORFs originating at each ancestral node, when normalized by branch length, decreases with increasing node-to-tip distance (Fig. 3C), reflecting dynamic turnover and reduced long-term retention of ncORFs. To assess whether individual ncORFs are constrained for coding, we analyzed codon substitution patterns within ncORFs since origination33 with PhyloCSF53. PhyloCSF estimates the log-likelihood ratio of a given alignment under coding versus non-coding models; positive values indicate stronger similarity to conserved coding sequences. We found 13.8% (1,608) of human and 32.0% (5,281) of mouse ncORFs have positive PhyloCSF scores (Fig. 3D), and 10.9% (1,264) of human and 22.6% (3,721) of mouse ncORFs even have PhyloCSF exceeding 10 (Table S6), providing strong evidence that these sequences may encode functional ncEPs.

Evolutionary dynamics can also provide insights into the function of ncORFs, as functionally constrained ncORFs are less likely to be lost over time. To quantify the conservation of each ncORF since its origination, we developed a local branch length score (BLS) (Fig. 4A), adapting the “global” BLS metric previously used to assess uORF conservation across full phylogenies48,54. Unlike global BLS, local BLS is designed to detect lineage-specific constraint, making it better suited for identifying functional ncORFs that may be restricted to certain evolutionary branches. Using a local BLS threshold of 0.9, we identified 16.3% (1,889) of human and 24.8% (4,087) of mouse ncORFs as exhibiting lineage-specific conservation (Figs. 4B & S9). Notably, these ncORFs are significantly more likely to show positive PhyloCSF scores (Fig. 4C), reflecting the effectiveness of local BLS. In total, we identified 681 human and 1,622 mouse ncORFs that are both lineage-specifically conserved and show evidence of coding potential, making them strong candidates for future functional validation. Together, these findings suggest that a subset of ncORFs may encode functional proteins rather than translational byproducts.

Lineage-specifically conserved ncORFs.

(A) Schematic illustration for BLS calculation. (B) Cladograms of vertebrates showing the number of lineage-specifically conserved ncORFs (local BLS > 0.9) at each ancestral node for humans and mice. (C) Distribution of PhyloCSF scores per codon for lineage-specifically conserved ncORFs compared to all other ncORFs. BLS, branch length score; ncORF, non-canonical open reading frame; ORF, open reading frame.

Expression of ncORFs is shaped by their evolutionary dynamics

Gene expression is a tightly regulated process, and non-functional products are unlikely to be robustly expressed. To assess the expression of ncORFs, we examined their translation across diverse tissues and cell lines (Table S7). On average, 71.9% of human and 48.5% of mouse ncORFs are translated with reads per kilobase per million mapped reads (RPKM) ≥ 1, which is comparable to or slightly exceeds the proportions observed for CDSs (Fig. S10). To investigate the influence of ncORF evolution on their expression, we analyzed mean translation levels of ncORFs across different samples. We found that ncORFs with higher local BLS tend to be more highly translated in both species (Fig. S11A), whereas coding potential showed only a minor influence (Fig. S11B). To disentangle the effects of evolutionary constraint and gene age, we stratified ncORFs by origination nodes and examined their translation levels in relation to local BLS and PhyloCSF. Although older ncORFs are generally more highly translated, those with higher local BLS scores consistently show elevated translation levels compared to their age-matched counterparts (Figs. 5A & S12A). This suggests that lineage-specific conservation is associated with enhanced and possibly optimized translation, potentially reflecting acquired functional roles. Conversely, the relationship between coding potential and translation is more nuanced: younger ncORFs with positive PhyloCSF scores tend to be more highly translated, while the opposite trend is observed in older ncORFs (Figs. 5B & S12B). These patterns point to complex interplays between evolutionary constraint, coding potential, and translational output, likely reflecting divergent selective pressures across different ncORF age groups.

Evolutionary dynamics of ncORF expression.

(A) Distribution of mean translation levels of human ncORFs grouped by their origin nodes and further stratified by local BLS. Statistical significance was assessed using Wilcoxon rank-sum tests. ***, P < 0.001; **, P < 0.01; *, P < 0.05. (B) Similar to (A) but ncORFs are further stratified by local BLS. (C) Relationship between ncORF tissue specificity at translation and transcription levels. Differences were determined with Wilcoxon signed-rank tests. (D-E) Similar to (A) and (B) but showing the distribution of tissue specificity at the translation level. CDS, coding sequence; ncORF, non-canonical open reading frame; ORF, open reading frame.

While the above analyses revealed the overall trends in ncORF translation, they overlooked the heterogeneity across tissues and cell types. We found that ncORF translation exhibits strong tissue specificity (Fig. S13), which may stem from either transcriptional bias in ncORF-containing genes or post-transcriptional regulation specific to ncORFs. To disentangle these factors, we examined RNA levels of ncORF-containing genes across different tissues55,56. Although transcriptional enrichment in testis was evident in both humans and mice (Fig. S14), testis-biased translation was observed only in mice (Fig. S13). Moreover, ncORF translation displayed significantly higher tissue specificity than the transcription of their host genes in both species (Fig. 5C), indicating a substantial post-transcriptional component to their regulation. We next assessed how evolutionary characteristics shape tissue specificity. Older ncORFs tend to have lower tissue specificity than younger ones, suggesting they are more broadly translated (Figs. 5D & S12C). Intriguingly, coding potential exhibits opposite effects depending on evolutionary age: younger ncORFs with higher coding potential are more broadly expressed, whereas older ones show increased tissue specificity (Figs. 5E & S12D). These patterns may help to explain the lower average translation levels observed for older ncORFs across tissues. A similar trend was also evident for ncORFs with high local constraint scores in mice (Fig. S12C). Collectively, these results support a model in which ncORFs initially evolve widespread expression to establish interactions and avoid loss57, but may later become functionally specialized, resulting in more restricted expression.

Extensive co-translation between ncORFs and canonical CDSs

Given our earlier observation that many ncEPs lack modular domains and may act via interactions with canonical proteins, we next explored the landscape of co-translation between ncORFs and CDSs across different tissues or cell types as we expect that interacting proteins tend to be translated simultaneously spatiotemporally. Using weighted gene co-expression network analysis (WGCNA)58, we identified 61,513 human and 55,449 mouse ncORF-CDS co-translated pairs. As expected, ncORFs in mRNAs tend to co-translate with the main CDS of the same gene, even after excluding CDS-overlapping ones (Fig. S15), consistent with previous experimental findings28. Among non-overlapping co-translating ncORFs, only 10 in humans were dORFs; all others in both species were uORFs. Of note, co-translation does not contradict the canonical model in which uORFs repress downstream CDS translation. Co-translation reflects concurrent ribosome occupancy driven by overall initiation rates at the 5’ end, whereas repression concerns the fraction of initiating ribosomes that reach the CDS. Thus, absolute translation of both ORFs may increase without altering the relative inhibitory effect of the uORF.

ncORF-CDS co-translation forms a highly modular bipartite network, in which one CDS may be linked to multiple ncORFs and vice versa (Fig. 6A-B). Network analysis revealed that hub CDS nodes—those with the most connections—in the largest modules are significantly enriched for functional gene categories (Fig. 6C-D). For instance, the largest module in humans is associated with chromatin-related proteins, while the second largest module in humans and the largest one in mice are linked to extracellular matrix organization. Notably, ancient ncORFs (those originating before the last common ancestor of mammals) are significantly more likely to co-translate with CDSs than younger ones (Fig. 6E), potentially reflecting a greater likelihood of acquiring functional interactions over evolutionary time. In summary, our analyses revealed that evolutionary dynamics of ncORFs have widespread impacts on their translation and potential function.

ncORF-CDS co-translation network.

(A) Bipartite network of ncORF-CDS co-translation in humans. (B) Bipartite network of ncORF-CDS co-translation in mice. ncORFs and CDSs are represented by different shapes and colored according to their cluster membership. Only the largest clusters were highlighted (top two in humans and top five in mice). (C) Top five enriched gene ontology terms for each of the two largest clusters in the human network shown in (A). (D) Top five enriched gene ontology terms for each of the three largest clusters in the mouse network shown in (B). (E) Proportion of ncORFs cotranslating with CDSs among ancient (pre-mammalian origin) versus younger (mammalian-specific) ncORFs. Differences were tested with Fisher’s exact tests. CDS, coding sequence; ncORF, non-canonical open reading frame; ORF, open reading frame.

Discussion

Reliable annotation of translated ncORFs is critical for understanding their functional roles and evolutionary significance. In this study, we developed a standardized pipeline for the annotation of translated ncORFs and applied it to both humans and mice due to the extensive publicly available Ribo-Seq datasets. Compared to previous efforts, our annotation framework offers several key improvements. First, we selected only high-quality libraries with sufficient read coverage and acceptable in-frame RPF rates. Our earlier work demonstrated that both library size and in-frame rates impact the reproducibility and accuracy of ncORF predictions20. This step helps mitigate false annotations arising from suboptimal data quality. Second, although many previously characterized functional ncEPs are associated with cancer-related ncORFs1, ncORFs may also have important functions under physiological conditions. Importantly, such ncORFs are more likely to be subject to evolutionary constraints. Therefore, we restricted our analysis to normal cells and tissues, allowing us to identify more biologically relevant and conserved ncORFs. Third, given the variable performance across ncORF detection methods and the low reproducibility of predictions between biological replicates20, we implemented stringent reproducibility criteria. Specifically, we retained only ncORFs detected by multiple methods and across independent datasets, enhancing the robustness of our final annotation. Finally, we applied a suite of well-validated filtering metrics from prior studies39-41 to minimize the inclusion of false positives, including those arising from translational noise59,60 or non-ribosomal contamination. As a result, our annotated ncORFs show significant enrichment for GENCODE ncORFs, especially those with MS evidence or reproducible detection across independent studies. These features collectively underscore the reliability and utility of our annotation for future functional and evolutionary studies of non-canonical translation.

However, our catalog may remain incomplete due to the uneven distribution of publicly available Ribo-seq datasets across tissues and cell types. Broader and more balanced tissue coverage will be necessary to generate a more comprehensive and less biased ncORF annotation, which would also enhance the cotranslation analysis. Additionally, our proteomic validation relies on previously compiled sets of MS-supported ORFs rather than de novo reanalysis of raw mass spectrometry datasets. Incorporation of candidate ncORFs into reference proteome databases prior to large-scale, standardized searches of public MS runs would substantially improve detection sensitivity and provide stronger evidence for their translational products45.

From the perspective of new gene evolution, ncORFs represent a transitional stage between non-coding DNA/RNA and protein-coding genes61,62. A major challenge in reclassifying ncORFs as bona fide coding genes lies in determining whether the putative proteins they encode are functionally relevant. Our analysis of ncEPs revealed that most lack recognizable protein domains and are enriched for IDRs. Therefore, we propose that many ncORFs may function in a non-autonomous manner by interacting with other proteins to exert their effects. This hypothesis aligns with prior studies showing physical interactions between ncEPs and known proteins28,29,34. Supporting this view, our WGCNA analysis uncovered widespread co-translation between ncORFs and canonical CDSs, suggesting potential interaction partners and functional contexts.

We also assessed whether ncORFs are subject to evolutionary constraints consistent with functional protein coding. Using both population-level and cross-species analyses, we found that a substantial fraction of ncORFs bear signatures of purifying selection, in line with previous studies33,63-66. Notably, their conservation profiles exhibit nucleotide-level features characteristic of canonical CDSs, supporting the interpretation that selection acts on the encoded peptide products rather than merely on translational activity per se. Furthermore, we reconstructed the evolutionary trajectories of thousands of ncORFs in both humans and mice and observed sustained coding potential since their origination. Complementing these findings, we identified a subset of ncORFs that preserve intact ORF structures across most descendant species, suggesting ongoing purifying selection against disabling mutations. These lineage-specifically conserved and potentially coding ncORFs likely contribute to clade-specific functional innovation and represent compelling candidates for future functional characterization.

We further explored how evolutionary history shapes ncORF translation across tissues or cell types. Our results indicate that older ncORFs are more broadly and highly translated and have established more extensive co-expression with canonical CDSs. Moreover, when comparing ncORFs of similar age, those that are lineage-specifically constrained consistently exhibit higher translation levels, reinforcing their likely functional importance. Interestingly, we found that younger ncORFs with strong evolutionary constraints tend to have lower tissue specificity in their translation, potentially reflecting a phase of widespread expression that facilitates functional recruitment. In contrast, older ncORFs with positive coding potential tend to display increased tissue specificity, implying functional specialization over evolutionary time. Collectively, these findings support a model in which ncORFs evolve from broadly expressed, weakly functional elements into specialized, conserved effectors— though further empirical work is needed to validate this hypothesis.

In conclusion, we provide a comprehensive annotation of translated ncORFs in humans and mice, offering a valuable resource for functional and evolutionary genomics. Our integrated analysis of sequence evolution and translational landscape sheds light on the evolutionary trajectory and functional integration of ncORFs. The consistency of these patterns in both species suggests that our findings may generalize to other mammals. However, realizing the full biological significance of ncORFs, particularly lineage-specific ones, will require experimental validation. Recent advances in high-throughput screening technologies19,29,32,67 offer a promising path forward. Lastly, although our study focused on humans and mice due to the availability of high-quality Ribo-Seq data, we anticipate that our annotation will serve as a training resource for machine learning models aimed at predicting translated ncORFs in non-model organisms, thus broadening our understanding of genome coding potential across species.

Methods

Quality control and preprocessing of Ribo-Seq libraries

Ribo-Seq libraries from normal human or mouse samples without genetic modifications or perturbations (Table S7) were downloaded from the European Nucleotide Archive (ENA) or the Short Read Archive (SRA). For ncORF annotation, libraries containing fewer than five million reads were excluded. Libraries derived from cancer tissues or cell lines were also removed based on sample metadata or the associated publications. Next, adapter sequences and any random nucleotide additions were trimmed using Cutadapt (v3.7)68 and reads shorter than 18 nucleotides were discarded. Remaining reads were aligned to the human or mouse reference genome (GRCh38 or GRCm39, ENSEMBL69 release 107) using the STAR aligner (v2.7.10a)70. To assess library quality, the proportion of in-frame ribosome-protected fragment (RPF) reads was calculated using the PSite package71. Libraries with ≥60% in-frame RPFs were classified as high quality. Only samples containing at least two high-quality libraries were retained for ncORF annotation (Table S1). For the expression analyses, all the downloaded Ribo-Seq libraries were used to cover more samples.

ncORF annotation

For each Ribo-Seq library, translated ORFs initiating with AUG, CUG, GUG, or UUG and encoding at least five amino acids were predicted using PRICE36, RiboCode37, and Ribo-TISH38, as described previously20. To define novel ORFs, annotated CDSs and their truncation or extension variants (i.e., ORFs with the same start or stop position as annotated CDSs) were excluded. To eliminate likely false positives resulting from non-ribosomal contamination or stochastic molecular errors, we calculated RRS40,41 and FLOSS39 of each ncORF. RRS measures the drop in RPF coverage after stop codon—a key signature of active translation beyond three-nucleotide periodicity40,41, whereas FLOSS assesses whether the RPF length distribution resembles that of canonical CDSs, helping to distinguish authentic ORFs from non-ribosomal signals39. In each library, uniquely mapped RPFs were assigned to the corresponding P-site, and genome-wide P-site coverage was computed with the PSite package. Similar to previous studies40,41, RRS was calculated as the sum of coverage across the five codons preceding the stop codon divided by the sum across the five trinucleotides following the stop codon. FLOSS was calculated by first deriving a reference RPF length distribution vector (Pref) using all RPFs uniquely mapped to annotated CDSs. For each ORF (ncORF or CDS), a read length frequency vector (PORF) was computed, and FLOSS was calculated as ∑|Pref − PORF| /2. The empirical upper bound for acceptable FLOSS was set using Tukey’s method (i.e., the third quartile + 1.5 × inter-quartile range of FLOSS values across all CDSs)39. ncORFs with fewer than 10 P-site-mapped RPFs, RRS ≤ 1, or FLOSS values exceeding the empirical upper bound were filtered out.

Integration of ncORF prediction across different libraries of the same species was nontrivial due to extensive overlap among ncORFs in genomic coordinates. To address this issue, we used a graph-based clustering approach. We created a graph where two ORF vertices were connected if they share the same start or stop codon. Then, each set of connected ncORFs in the graph represents a cluster of overlapping ncORFs. To ensure reproducibility, only clusters identified by ≥2 methods in ≥2 libraries were retained, yielding 13,071 clusters in humans and 17,674 clusters in mice. To ensure compatibility with the latest reference gene annotations (ENSEMBL release 111), we removed any ncORFs that were newly classified as annotated CDSs. A manual inspection of RPF signals using the GWIPS-viz Ribo-Seq track72 in the UCSC Genome Browser73 revealed additional spurious ncORFs overlapping in-frame with CDSs from other transcripts. These were likely misidentified due to shared translation signals and were conservatively excluded using custom scripts, leaving 11,623 clusters in humans and 16,485 in mice.

To facilitate downstream analyses, a representative ncORF was selected for each cluster using a two-step strategy. First, a representative transcript was selected by prioritizing those with matched annotation from NCBI and EMBL-EBI (MANE)42 (for humans) or APPRIS principal74; otherwise, the longest transcript was chosen. Then, among ncORFs within the representative transcript, a single ncORF was selected based on start codon hierarchy (AUG > CUG/GUG > UUG) following empirical evidence28,36. If multiple ncORFs shared the same preferred start codon, the longer one was chosen.

For validation, our ncORFs were compared with previously annotated GENCODE ncORFs2 and those with MS evidence compiled previously20. Tier annotations of GENCODE ncORFs were obtained from a recent study45. Known functional ncORFs and their genomic coordinates were manually curated from published literature (Table S4). To account for possible proteoforms arising from alternative initiation and splicing, we considered an ncORF to be recovered if either its start or stop codon was identical to any ncORF in our annotation.

ncORF sequence analyses

The secondary structure surrounding each ORF (ncORF or CDS) start codon was predicted with RNAfold75 on a 150 nt window extracted from the transcript in which the ncORF resides starting from the 30th nucleotide upstream of the start codon. For each position within this window, the overall pairing probability was calculated by dividing the number of ORFs with a base-paired residue at that position by the total number of ORFs analyzed. The tRNA adaptation index and effective number of codons for ncORFs and CDSs were calculated using the “cubar” R package76. tRNA gene annotations in humans and mice were obtained from GtRNAdb (release 22)77.

Putative ncEPs analyses

Known domains within ncEPs were identified using InterProScan (v5.64-96.0)78 with default parameters. Repetitive sequence annotations were downloaded from the UCSC Table Browser. BEDTools (v2.31.1)79 was used to identify TE-overlapping ncORFs. IDRs were predicted with IUPred380 with analysis type set to “short disorders” and smoothing disabled. For each ORF, residues with disorder tendency score > 0.5 were considered disordered, and the IDR fraction was calculated as the proportion of disordered residues relative to the ORF length.

Gnocchi Score and PhyloP of ncORFs and CDSs

Genome-wide Gnocchi scores computed using 1 kb sliding windows (with 100bp step size) were downloaded from the gnomAD database50. For each base pair, the Gnocchi score was defined as the average across the ten overlapping 1kb windows that covered that position using BEDTools. The final Gnocchi score for each ORF (ncORF or CDS) was obtained by averaging the scores of all base pairs within the ORF.

Base-wise PhyloP scores computed across 441 mammals or 239 primates51 were downloaded from the UCSC Genome Browser. For each ncORF or CDS, PhyloP values were extracted for a ±30 bp window surrounding the start and stop codons using custom scripts. These values were then averaged over different ORFs to generate metagene profiles for visualization. To evaluate three-nucleotide periodicity, the metagene profiles were subjected to autocorrelation analysis using the “ac.test” function from the “testcorr” R package81. Statistical significance was assessed using the robust t~ statistics.

ncORF origination node and mode

Whole genome alignments and corresponding species trees were downloaded from the UCSC Genome Browser (GRCh38 100-way for humans and GRCm39 35-way for mice). The origination node and mode for each ncORF were inferred following previous approaches33,34. For each ORF, the sequence from the reference genome and its orthologous regions in other species were extracted. Orthologous sequences covering less than 50% of the ncORF length relative to the reference species were excluded. The remaining sequences were realigned with MAFFT (v7.520)82 with default parameters. The corresponding gene tree was derived by pruning the species tree using Gotree (v0.4.3)83 with default settings. Ancestral sequence reconstruction was performed using FastML (v3.11)84 with parameters “--seqType nuc --SubMatrix HKY --OptimizeBL no --indelReconstruction ML”. An ORF structure was considered intact in an orthologous or ancestral sequence if one of the first three codons was a start codon and at least 70% of the 5’ region of the reference ncORF was free of in-frame stop codons. The origination node was defined as the most ancient ancestral node that retained an intact ORF structure. An ncORF was classified as de novo if the origination node had an ancestor covering at least 50% of the ncORF sequence but lacking an intact ncORF structure. Otherwise, the origination mode was classified as uncertain. The distance from an ancestral node to the reference species (tip node) was calculated by summing branch lengths along the connecting path using the ape R package (v5.8)85.

ncORF BLS and PhyloCSF

For each ncORF, global BLS was calculated as the ratio of the total branch length of the subtree connecting all leaves with intact ORF structures to the total branch length of the full species tree as in our previous study48 with custom scripts. Local BLS was calculated similarly, but the denominator was replaced by the total branch length of the subtree defined by the origination node and all of its descendant nodes (including leaves). To evaluate coding potential, the alignment of all species descending from the origination node was converted into Multiple Alignment Format after removing the stop codon. PhyloCSF scores were then computed using the “score-msa” module of PhyloCSF++86.

ncORF expression analysis

To quantify translation levels of ncORFs and CDSs in Ribo-Seq libraries, we assigned uniquely mapped reads to corresponding P-site positions with the PSite package. The read count for each ORF was calculated as the total number of RPFs whose P-sites fell within the ORF. To reduce bias caused by RPF enrichment at ORF termini, reads mapping to start or stop codons were excluded. For CDSs, only the count of the representative transcript of each gene as defined above was used for downstream analyses. For samples with multiple libraries, read counts were aggregated prior to normalization. Sample-specific size factors were estimated using CDS counts and applied to normalize both ncORF and CDS counts across samples using DESeq287. RPKM for an ORF was calculated as normalized read counts / ORF length (nt) / library size × 109, where library size was the median of total read counts across samples to ensure cross-sample comparability. An ORF (ncORF or CDS) was considered translated if its RPKM was ≥ 1. To evaluate tissue specificity of ncORF translation, the normalized RPF count Xk of an ORF in sample k was log-transformed as xk = log(Xk + 1). Then, tissue specificity index88 was calculated as τ=∑kn(1−x~k)/(n−1), where x~k=xk/max1≤j≤nxj and n is the total number of samples.

To quantify RNA levels, raw reads of deep transcriptome profiling across different human or mouse tissues from previous studies55,56 were downloaded from SRA. Gene expression levels (transcript per million, TPM) were estimated using salmon (v1.10.3)89 with parameters “--seqBias --gcBias –posBias -l A”.

WGCNA analysis58 of ncORF and CDS translation was performed using the PyWGCNA package (v2.0.5)90 with default settings. Only ORFs with normalized RPF counts ≥ 10 in more than half of the samples were included. The soft threshold was determined automatically by PyWGCNA. Due to differences in soft thresholds, two ORFs were considered cotranslated if their topological overlap > 0.1 (human) or > 0.01 (mouse).58 To test whether co-translation is enriched between ncORFs and CDSs from the same transcript, gene identities of CDSs were randomly permuted among all co-translated ncORF–CDS pairs. This procedure was repeated 1,000 times, and the empirical P value was calculated as (m + 1) / 1,000, where m is the number of permutations in which the number of same-gene pairs was greater than or equal to the observed count. The bipartite network of ncORF-CDS co-translation was clustered using the Leiden algorithm91 implemented in igraph (v2.1.3)92 by optimizing modularity with a resolution of 0.2. The top 20% of nodes by betweenness centrality were designated as hub nodes. Gene ontology analysis of hub genes in the largest modules was conducted using ClusterProfiler (v4.12)93.

eLife Assessment

This study presents a large, systematically curated catalog of non-canonical open reading frames (ncORFs) in human and mouse through the reanalysis of nearly 400 Ribo-seq datasets using a standardized pipeline; the resulting atlas consolidates ncORF annotations across tissues and provides a valuable resource for investigating non-canonical translation and ORF emergence. The main conclusions are supported by consistent data processing and multiple computational measures of translation and conservation. While the pipeline is transparent and technically robust, some analytical criteria and dataset limitations could be described more explicitly, and several downstream conclusions would benefit from more cautious interpretation, some evolutionary inferences are primarily correlative; dataset heterogeneity, uneven tissue representation, and limited experimental validation also constrain the strength of a subset of the findings. Overall, the evidence is solid, and the resource is likely to be broadly beneficial to the community.

Reviewer #1 (Public review):

This work compiles a comprehensive atlas of ncORFs across mammalian tissues and cell types, derived from reanalysis of ~400 public ribosome profiling datasets. The authors then evaluate cross-species conservation and functional signatures, proposing that evolutionarily ancient ncORFs tend to have higher translation potential, stronger expression, and closer relationships with canonical coding sequences.

Strengths:

In general, the study provides a large-scale and timely resource of annotated ncORFs, which could be broadly useful for the community. The authors collected ~400 public ribosome profiling datasets for annotations of ncORFs, which, to my best knowledge, is the largest collection of data for such purpose. The catalog could facilitate future investigations into ncORF biology and broaden understanding of the coding potential of the "non-coding" genome.

Weaknesses:

Based on the ncORF catalog, some of the analyses were not properly done. Some of the results are descriptive.

(1) Bias and representations of data source. Public ribo-seq datasets are unevenly distributed across tissues and cell lines, raising concerns about heterogeneity and underrepresentation of certain contexts. This may limit the generalizability of the catalog.

(2) The discussion on modular domains of ncORFs is unclear, and the claim that they may originate via TE-related mechanisms is not well supported. Stronger evidence or clearer reasoning is needed.

(3) The conservation comparisons are not fully convincing. Figure S7 shows only mild differences between ncORFs and CDS, and statistical significance is not clearly demonstrated. Comparisons with other non-coding RNAs should be added, and overlapping sequences between ncORFs and CDS should be excluded to avoid bias.

(4) Figure 3 indicates that some ncORFs are subject to evolutionary constraints. This is not surprising. The authors should provide further analyses on more detailed features of these "conserved" ncORFs vs. the "non-conserved" ones. Some pretty informative works have been done in drosophila, worms, mouse, and human. Figure 3 suggests some ncORFs are under evolutionary constraint, but this is not unexpected. More granular analyses contrasting "conserved" versus "non-conserved" ncORFs would be informative. In fact, small ORFs, especially uORFs, have been extensively studied, for their functions and corss-species conservations. The authors should explicitly show what is new here in their analyses.

(5) Translation levels are reported using RPF counts. However, translation efficiency (normalized by RNA expression) is a more appropriate measure to account for expression heterogeneity.

(6) The correlation analyses between ncORF translation levels and PhyloCSF are confusing and largely descriptive. These sections need sharper framing and clearer conclusions.

(7) Public ribo-seq datasets, generated by different research labs, are known for their strong batch effects. Representations of tissues and cells are also very unbalanced. Therefore, the co-translation analysis between ncORFs and canonical CDS is not well controlled. This should be done by referring to a recent large-scale ribo-seq meta-analysis (Nat Biotechnol. 2025. doi: 10.1038/s41587-025-02718-5).

Comments on revisions:

The authors have made efforts to address most of the previous concerns, and several points have been clarified or improved in the revision. However, in a number of cases, the responses rely more on acknowledgment and reframing rather than substantive analytical strengthening. Overall, the manuscript is improved, particularly in terms of clarity, transparency, and positioning of claims. I support its publication and look forward to seeing how the field engages with and discusses these claims.

Reviewer #2 (Public review):

Summary:

Chang et al. attempted to analyze a large number of ribo-seq datasets through a standardized pipeline, identifying novel non-canonical ORFs and elucidating their evolutionary and expression characteristics.

Strengths:

(1) The datasets analyzed by the authors are sufficiently comprehensive, and the use of standardized pipelines ensures excellent analytical consistency.

(2) Their analyses of ORF evolution and co-expression further deepen our understanding of these ORFs.

Weaknesses:

(1) The authors primarily conducted analyses through bioinformatics, lacking sufficient wet-lab experimental evidence.

(2) Some analytical methods and standards were not clearly presented in the manuscript.

Author response:

The following is the authors’ response to the previous reviews.

Public Reviews:

Reviewer #1 (Public review):

This work compiles a comprehensive atlas of ncORFs across mammalian tissues and cell types, derived from reanalysis of ~400 public ribosome profiling datasets. The authors then evaluate cross-species conservation and functional signatures, proposing that evolutionarily ancient ncORFs tend to have higher translation potential, stronger expression, and closer relationships with canonical coding sequences.

Strengths:

In general, the study provides a large-scale and timely resource of annotated ncORFs, which could be broadly useful for the community. The authors collected ~400 public ribosome profiling datasets for annotations of ncORFs, which, to my best knowledge, is the largest collection of data for such purpose. The catalog could facilitate future investigations into ncORF biology and broaden understanding of the coding potential of the "non-coding" genome.

Weaknesses:

Based on the ncORF catalog, some of the analyses were not properly done. Some of the results are descriptive.

(1) Bias and representations of data source. Public ribo-seq datasets are unevenly distributed across tissues and cell lines, raising concerns about heterogeneity and underrepresentation of certain contexts. This may limit the generalizability of the catalog.

(2) The discussion on modular domains of ncORFs is unclear, and the claim that they may originate via TErelated mechanisms is not well supported. Stronger evidence or clearer reasoning is needed.

(3) The conservation comparisons are not fully convincing. Figure S7 shows only mild differences between ncORFs and CDS, and statistical significance is not clearly demonstrated. Comparisons with other noncoding RNAs should be added, and overlapping sequences between ncORFs and CDS should be excluded to avoid bias.

(4) Figure 3 indicates that some ncORFs are subject to evolutionary constraints. This is not surprising. The authors should provide further analyses on more detailed features of these "conserved" ncORFs vs. the "non-conserved" ones. Some pretty informative works have been done in drosophila, worms, mouse, and human. Figure 3 suggests some ncORFs are under evolutionary constraint, but this is not unexpected. More granular analyses contrasting "conserved" versus "non-conserved" ncORFs would be informative. In fact, small ORFs, especially uORFs, have been extensively studied, for their functions and corss-species conservations. The authors should explicitly show what is new here in their analyses.

(5) Translation levels are reported using RPF counts. However, translation efficiency (normalized by RNA expression) is a more appropriate measure to account for expression heterogeneity.

(6) The correlation analyses between ncORF translation levels and PhyloCSF are confusing and largely descriptive. These sections need sharper framing and clearer conclusions.

(7) Public ribo-seq datasets, generated by different research labs, are known for their strong batch effects. Representations of tissues and cells are also very unbalanced. Therefore, the co-translation analysis between ncORFs and canonical CDS is not well controlled. This should be done by referring to a recent large-scale ribo-seq meta-analysis (Nat Biotechnol. 2025. doi: 10.1038/s41587-025-02718-5).

Comments on revisions:

The authors have made efforts to address most of the previous concerns, and several points have been clarified or improved in the revision. However, in a number of cases, the responses rely more on acknowledgment and reframing rather than substantive analytical strengthening. Overall, the manuscript is improved, particularly in terms of clarity, transparency, and positioning of claims. I support its publication and look forward to seeing how the field engages with and discusses these claims.

We thank Reviewer #1 for the constructive comments and recognition of the value of our ncORF atlas. We have addressed the key concerns by strengthening the analyses and clarifying the framing and limitations of our conclusions. We appreciate the reviewer’s support for publication and believe these revisions have further improved the manuscript.

Reviewer #2 (Public review):

Summary:

Chang et al. attempted to analyze a large number of ribo-seq datasets through a standardized pipeline, identifying novel non-canonical ORFs and elucidating their evolutionary and expression characteristics.

Strengths:

(1) The datasets analyzed by the authors are sufficiently comprehensive, and the use of standardized pipelines ensures excellent analytical consistency.

(2) Their analyses of ORF evolution and co-expression further deepen our understanding of these ORFs.

Weaknesses:

(1) The authors primarily conducted analyses through bioinformatics, lacking sufficient wet-lab experimental evidence.

(2) Some analytical methods and standards were not clearly presented in the manuscript.

We thank Reviewer #2 for the positive assessment of our comprehensive dataset and standardized analytical framework. We have clarified the analytical methods and criteria throughout the manuscript and better defined the scope and limitations of our bioinformatics-based analyses. We appreciate the reviewer’s constructive suggestions, which have helped improve the clarity and rigor of the manuscript.

Recommendations for the authors:

Reviewing Editor:

We have evaluated the revision together with the reviewers' second-round assessments and your responses. The reviewers agree that the manuscript has improved in clarity and that the standardized integration of large-scale Ribo-seq datasets provides a valuable resource for the field. However, several important concerns remain insufficiently resolved. In multiple cases, the revision relies primarily on acknowledgment or reframing of limitations rather than additional analyses or clearer methodological justification, leaving the evidential support for several conclusions incomplete.

Because the study is entirely computational, all analytical procedures, criteria, and thresholds should be explicitly defined and adequately justified to meet the expected standard of technical rigor. In particular, key components of the analytical framework require clearer description, including the definitions and criteria used for co-translation and ncORF classification. The limitations of the dataset should also be discussed more explicitly, especially regarding the heterogeneity of public Ribo-seq datasets, technical factors influencing detection sensitivity, and the interpretation of variable detection frequencies across samples.

In addition, several conclusions remain largely descriptive, and the distinction between novel findings and confirmation of previous observations should be clarified more carefully. Conclusions should be framed strictly within the limits of the presented data and positioned appropriately relative to prior work, with suitable citation to avoid overstating novelty.

We therefore ask the authors to refine the technical descriptions, ensure that all methods and analytical criteria are presented unambiguously, and expand the Discussion to clearly articulate the limitations of the dataset and analysis. The conclusions should also be revised to reflect an appropriately cautious interpretation of the findings.

The primary strength of this study lies in the scale and standardization of the dataset as a community resource. Given this substantial resource value, we believe the manuscript could become suitable for publication provided that the issues outlined above are addressed clearly and transparently. With these revisions, the work will provide a useful foundation for future studies in this area.

We therefore invite you to submit a final revised version addressing the points described above.

We thank the Editor for the careful assessment and constructive guidance. In the final revision, we have clarified all key methodological definitions and analytical criteria, expanded the Discussion of dataset and detection limitations, and revised the conclusions to avoid overstatement. We believe these changes improve the technical rigor, transparency, and overall value of the manuscript as a community resource.

Reviewer #1 (Recommendations for the authors):

The authors have made efforts to address most of the previous concerns, and several points have been clarified or improved in the revision. However, in a number of cases, the responses rely more on acknowledgment and reframing rather than substantive analytical strengthening. Overall, the manuscript is improved, particularly in terms of clarity, transparency, and positioning of claims. I support its publication and look forward to seeing how the field engages with and discusses these claims.

We appreciate the reviewer’s support for publication.

Reviewer #2 (Recommendations for the authors):

While the authors have made commendable efforts to revise the manuscript and address previous concerns, the revised version still falls short of fully resolving several critical issues regarding data interpretation and methodological transparency. I recommend the following modifications before the manuscript can be considered for publication:

(1) Although the authors have annotated the detection frequencies of individual sORFs in the revised

Supplementary Table 3 and Fig. S1B, the biological and technical implications of these data require further clarification:

(a) As the data demonstrates, even the most abundant sORFs were detected in no more than half of the samples. If the authors attribute this low detection rate to technical limitations (e.g., batch effects, sequencing depth, or threshold stringency), this must be explicitly discussed and annotated in the main text to prevent readers from misinterpreting this as low biological penetrance.

We thank the reviewer for this constructive comment. We have added a paragraph to the Discussion explicitly addressing the potential technical factors underlying the variable detection frequencies to avoid overinterpreting these frequencies as biological penetrance.

(b) The authors did not fully address my previous query regarding tissue specificity. Given the diverse and complex origins of the analyzed ribo-seq datasets, it is crucial to know whether any of these sORFs are tissue-specifically translated. The authors should analyze and state whether certain sORFs are exclusively detected in specific sample categories (tissues/organs), and whether this translation pattern aligns with the tissue-specific expression of their corresponding host transcripts.

We thank the reviewer for raising this important point. Strict tissue-exclusive translation is difficult to establish from heterogeneous public Ribo-seq datasets, as gene expression is inherently stochastic and most genes have some probability of being expressed across tissues, although expression levels may vary substantially between tissues. Moreover, failure to detect an ncORF in a given tissue may reflect low expression or insufficient sequencing depth rather than true biological absence. We therefore avoid making definitive claims about tissue-specific translation and instead quantify variation in ncORF expression across tissues using a tissue specificity index.

(2) Echoing the concerns raised by Reviewer #1, I remain concerned that the observed uORF-CDS cotranslation might be a computational artifact or false positive. The current Methods section lacks sufficient detail on how "co-translation" was strictly defined and quantified. I strongly recommend that the authors include a schematic diagram (e.g., in Figure 6 or supplementary figures) that explicitly details their analytical strategy, statistical thresholds, and evaluation criteria for defining co-translation. As was pointed out, it is well-established that uORFs typically exert an inhibitory effect on the translation of the main CDS. To validate the accuracy and robustness of their analytical pipeline, the authors should use their collected dataset to demonstrate the prevalence and nature of this canonical inhibitory phenomenon. Successfully capturing this expected repression would serve as a crucial positive control for their methodology. While the authors provided a theoretically acceptable mechanistic model in the text to reconcile co-translation with uORF-mediated repression, this hypothesis currently lacks literature support. The authors must cite relevant prior studies that support this specific regulatory dynamic to strengthen their argument.

We thank the reviewer for this important comment. We have further clarified the definition, detection criteria, and statistical framework for co-translation in the Methods, and have made the complete analysis code publicly available to facilitate reproducibility and independent evaluation. We have also added relevant literature supporting the proposed regulatory interpretation clarifying the relationship between our observations and the established inhibitory effects of uORFs.

Version 3

Abstract

Non-canonical open reading frames (ncORFs) are pervasive within transcripts annotated as “non-coding” or “untranslated regions” of mRNAs, yet their landscape under normal physiological conditions remains to be fully resolved, particularly outside humans. Here we applied a stringent and standardized pipeline to hundreds of high-quality ribosome profiling libraries from normal mammalian tissues and cell types, identifying 11,623 human and 16,485 mouse ncORFs. Evolutionary analyses revealed that thousands of ncORFs are subject to coding constraint and exhibit lineage-specific conservation, underscoring their functional potential. Ancient ncORFs are preferentially highly translated, broadly expressed, and enriched for lineage-specific conservation. Co-expression patterns further indicate that many ncORFs, especially ancient ones, are cotranslated with canonical coding sequences, consistent with functions mediated through protein–protein interactions. Together, these findings establish a comprehensive atlas of mammalian ncORFs and provide fundamental insights into their evolutionary dynamics and functional integration within the proteome.

Introduction

The systematic identification of functional genomic elements represents a foundational objective in the field of genomics. Among these elements, non-canonical open reading frames (ncORFs)—which are open reading frames (ORFs) located within long non-coding RNAs (lncRNAs) or untranslated regions (UTRs) of mRNAs—have attracted considerable attention in recent years1–5. While certain ncORF subtypes, such as upstream ORFs (uORFs) located in 5’ UTRs and downstream ORFs (dORFs) in 3’ UTRs, have been extensively characterized for their ability to regulate mRNA translation and stability in cis, a growing body of evidence demonstrates that ncORFs can also give rise to functional proteins that engage in diverse cellular pathways1,2,6,7. As functional ncORFs continue to be discovered across multiple eukaryotes8–17, systematic elucidation of their composition and function promises to uncover an additional layer of regulatory complexity and to refine our conception of the so-called “non-coding” genome.

Although ncORFs may play important biological roles, the number of translated ncORFs in the human genome remains uncertain. Detection by mass spectrometry (MS)-based proteomics is challenging, likely due to their short lengths and low abundances18. Ribosome profiling (Ribo-Seq) offers greater sensitivity for identifying translation events18,19, but downstream computational pipelines often produce highly divergent ncORF sets 20. Consequently, published estimates of ncORF abundance span several orders of magnitude—ranging from thousands to millions 15–22. To establish a standardized reference, the GENCODE consortium integrated data from seven distinct studies to produce a catalog of 7,264 human ncORFs (hereafter GENCODE ncORFs)2; however, the heterogeneity of sample sources and prediction algorithms likely introduced systematic bias. This catalog is also restricted to AUG-initiating ORFs, thereby overlooking those with near-cognate start codons. A later effort generated a high-resolution human translatome map and identified 7,767 ncORFs with a unified pipeline22. Nevertheless, this conservative annotation disproportionately under-represents coding sequence (CDS)-overlapping ncORFs. Indeed, later estimation suggests that between 10,500 to 22,500 ncORFs should exist in the human genome18. Furthermore, many ncORFs were reported from studies of cancer cells1,13,23,24, where extensive transcriptome reprogramming alters the translational landscape25,26. While paired healthy tissues were also analyzed in these studies, the landscape of ncORFs translated under normal physiological conditions is not yet fully resolved.

While a limited subset of ncORFs has been shown to encode signaling molecules, cofactors, membrane proteins, or subunits of large protein complexes1,27,28, functions of most ncORF-encoded proteins (ncEPs) are largely uncharacterized. Advances in high-throughput screen technologies have enabled the systematic interrogation of functional ncEPs23,28–32. Nevertheless, there is minimal concordance among hits from different studies3. Evolutionary analyses can offer insights into potential functionality of these translation products and mechanisms of origination33–35, yet how ncORF translation has been remodeled to integrate into proteome since their evolutionary origins is still poorly understood. Consequently, the extent to which translated ncORFs encode functional proteins and the evolutionary dynamics of their integration into the existing proteome are largely unresolved. Moreover, because most ncORF studies have focused on humans, it is unclear whether ncORFs in other species share similar sequence features and expression patterns.

To address these gaps, we generated a standardized annotation of translated ncORFs by uniformly reprocessing ~400 high-quality ribosome profiling libraries from normal mammalian tissues and cell types. This stringent approach yielded 11,623 human and 16,485 mouse ncORFs, consistently recovered across diverse samples and prediction strategies. Comparative and functional genomic analyses revealed that thousands of these ncORFs are under evolutionary constraint and exhibit distinct translation profiles. The resulting dataset provides a robust, comprehensive resource that will facilitate future investigations into the biological roles of ncORFs across cellular and organismal contexts.

Results

Annotation of ncORFs translated in normal mammalian cells

We developed a standardized pipeline to systematically annotate ncORFs translated in normal human and mouse cells (Fig. 1A). We first curated publicly available Ribo-Seq libraries, excluding those from pathological or genetically modified samples. Since both library size and the proportion of in-frame ribosome-protected fragments (RPFs) affect the sensitivity and accuracy of ncORF prediction20, we filtered out libraries containing fewer than 5 million RPFs or with in-frame rates below 60%. After this quality control step, we retained 174 human and 209 mouse high-quality Ribo-Seq libraries, spanning a range of tissues and cell types (Fig. S1A; Table S1). We predicted translated ORFs that begin with NUG (where N stands for any nucleotide) start codons and encode at least five amino acids using three tools—PRICE36, RiboCode37 and Ribo-TISH38—which we previously benchmarked for optimal performance20. After excluding annotated CDSs and their extension or truncation variants arising from alternative initiation sites, we obtained a preliminary list of ncORFs in each library.

Annotation of ncORFs in humans and mice.

(A) Overview of the integrative pipeline for ncORF annotation in human and mouse genomes. (B-C) RPF coverage plot for a representative uORF (B) and a lncORF (C), visualized using the Ribo-Seq signal track from GWIPS-viz72. ncORFs were shown in yellow and CDSs in cyan. Dashed lines indicate identical genomic regions shared across transcript isoforms. (D-E) Proportion of GENCODE ncORFs rediscovered in this study. GENCODE ncORFs were stratified either by reproducibility across independent studies (D) or by tiered translation evidence strength (E). (F) Known ncORFs that were experimentally characterized in earlier studies and independently rediscovered in this study from humans and mice. CDS, coding sequence; FLOSS, fragment length organization similarity score; MANE, matched annotation from NCBI and EBI; ncORF, non-canonical open reading frame; ORF, open reading frame; RPF, ribosome-protected fragment.

To generate reliable ncORF annotations, we applied stringent filtering and integration criteria. We removed potential false positives caused by non-ribosomal contamination or translational noise based on metrics including Fragment Length Organization Similarity Score (FLOSS)39, Ribosomal Release Score (RRS)40,41 and the number of RPFs within an ORF by P-site. We then aggregated the remaining ORFs to identify those consistently detected across multiple methods and libraries in human and mouse datasets, respectively (Fig. 1A). In many cases, two ORFs predicted from different libraries or methods may share the same start or stop codon due to the alternative translation initiation or splicing, resulting in groups of overlapping ncORFs. For example, 11.1% (805) of GENCODE ncORFs overlap with at least one other ORF, forming 329 overlapping groups (Table S2). This overlap complicates the identification of reproducible ORFs, as ORFs within a group may represent distinct proteoforms of a protein. To address this issue, we collapsed each group of ncORFs sharing the same start or stop codon into a single cluster, and retained only those clusters supported by at least two independent methods and observed in more than one library. For each retained cluster, we selected a representative ORF by prioritizing transcript annotation42,43, start codon type, and ORF length to facilitate downstream analyses (Fig. 1A).

This approach yielded 12,469 translated ncORFs in humans and 16,960 in mice. However, manual inspection revealed a subset of spurious ncORFs that are not truncation or extension variants of annotated CDSs, but instead partially overlap with annotated CDSs from alternative transcript isoforms in frame (Figs. 1B & C). For instance, in gene AGTPBP1, the CDS of transcript ENST00000357081 begins downstream of ENST00000628899 CDS due to the inclusion of an alternative exon. As a result, the 5’ UTR of ENST00000357081 appears to contain a translated uORF (Fig. 1B), which overlaps in-frame with the CDS of ENST00000628899 and may be mistakenly identified as a translated ncORF owing to active translation of the latter. A similar case is seen with the lncRNA ORF (lncORF) in ENST00000462211, which overlaps in-frame with the CDS of ENST00000328767 from the TMTC4 gene and may thus be erroneously identified as a translated ncORF (Fig. 1C). To avoid such artifacts, we implemented an additional filtering step to exclude these overlapping cases and left with 11,623 and 16,485 high-confidence ncORFs in humans and mice, respectively (Table S3). Notably, while more than half ncORFs were detected in two to five libraries, 22.3–26.3% were identified in more than 10 libraries, indicating a substantial subset with robust and recurrent translation signals suitable for future experimental validation (Fig. S1B).

Annotated ncORFs are supported by independent evidence

To assess whether the annotated ncORFs are supported by independent evidence, we compared them with a previously compiled catalog of ncORFs identified using mass spectrometry (MS)-based proteomic data20. For the 3,494 MS-supported ncORFs in humans and 873 in mice, 44.5% and 67.6%, respectively, were also predicted as translated ncORFs in this study (Fig. S2). We further evaluated our annotation against the GENCODE ncORFs2, a widely used community resource. Of the 7,264 GENCODE ncORFs, 6,974 are located in fully annotated transcripts that were included in our analysis. Among them, 2,997 were reported by at least two independent studies (hereafter “phase 1”), while the remaining were only reported in a single study (“single”)2. We found that 75.6% of phase 1 ncORFs were independently recovered in our annotation, significantly more than the single-study ncORFs (Fig. 1D; P = 6.4×10−307, Fisher’s exact test).

A recent study conducted a comprehensive pan-proteome analysis to provide high-confidence peptide evidence for GENCODE ncORFs44. In this study, ncORFs were categorized into four tiers: tier 1 and 2 ncORFs had support from both MS and Ribo-Seq data; tier 3 had MS evidence only; and tier 4 had Ribo-Seq evidence only45. Notably, our annotation was significantly enriched for tier 1 or 2 ncORFs (P = 2.0×10−64, Fisher’s exact test), with 446 and 661 of them detected, respectively (Fig. 1E). Combining MS-supported ncORFs from our previous compilation and those from GENCODE ncORFs, 2,088 (18.0%) human ncORFs identified in this study were supported by MS evidence. Despite this overlap, the majority (65.2%) of human ncORFs in our annotation are not included in either of the two catalogs. This highlights both the complexity of the non-canonical translatome and the critical role of tissue diversity in comprehensive ncORF annotation.

To further validate our findings, we manually curated ncORFs with known functions from the literature. We identified 17 ncORFs in our annotation that have been experimentally confirmed to be translated—either through reporter assays or antibody detection—and shown to produce functional proteins (Fig. 1F; Table S4). Notably, 16 of these validated ncORFs are from humans, and only 12 are included in the GENCODE catalog. This analysis further underscores the accuracy and functional relevance of our annotations. Together, these results indicate that our annotation pipeline generates a robust and reliable catalog of translated ncORFs in mammals.

ncORFs exhibit signatures of translation and encode putative proteins lacking known domains

We classified ORFs into categories based on community recommendations2 and found that the distribution of ORF types closely aligns with previous estimates18. In both human and mouse, uORFs, upstream overlapping ORFs (uoORFs), and lncORFs are more prevalent than other types, whereas internal ORFs (iORFs), downstream overlapping ORFs (doORFs), and dORFs are relatively rare (Fig. 2A). Although most ncORFs are short, encoding fewer than 100 amino acids (Fig. 2B), their sequences show clear signatures of translation similar to CDSs. These include preferential usage of AUG start codons (Fig. S3A), enrichment toward 5’ transcript ends (Fig. S3B), and reduced RNA secondary structure46 (Fig. S3C) and optimized Kozak contexts47,48 around start codons, especially in non-AUG ncORFs (Figs. S3D & S4). Despite these similarities, ncORFs differ markedly from CDSs in both codon and amino acid usage (Figs. S5A-B). This divergence is further supported by their lower tRNA adaptation index values and higher effective number of codons, indicating less optimized codon usage (Figs. S5C-D). Together, these findings show that ncORFs share core sequence and structural features with canonical CDSs, yet maintain distinct codon and amino acid usage profiles.

Sequence features of ncORFs and putative ncEPs in humans and mice.

(A) Number of ncORFs belonging to different categories. (B) Distribution of ncORF lengths. (C) Proportion of ncEPs that contain known Pfam domains. (D) Number of ncORFs with Pfam domains among lncORFs and other ncORF categories. (E) Distribution of the proportion of intrinsically disordered residues in CDS- and ncORF-encoded proteins. (F) Most frequently identified Pfam domains among human and mouse ncORFs. (G) Proportion of ncORFs overlapping with TEs, stratified by presence or absence of Pfam domains. Differences were assessed using Wilcoxon rank-sum tests. CDS, coding sequence; ncORF, non-canonical open reading frame; ncEP, ncORF-encoded protein; TE, transposable element.

Although ncORFs can encode putative proteins, the functional significance of ncEPs remains poorly understood. As an initial step toward functional inference, we systematically searched for annotated protein domains within ncEP sequences, reasoning that the presence of conserved domains would suggest potential biochemical activity. Overall, only 1.29% (150) of human and 0.78% (129) of mouse ncEPs contained at least one known domain (Fig. 2C; Table S5), presumably due to their short lengths relative to canonical proteins. Consistently, domain-containing ncORFs are significantly longer than those without annotated domains (Fig. S6) and are predominantly classified as lncORFs (Fig. 2D). Unlike canonical proteins, which have relatively low intrinsically disordered region (IDR) content (median ± standard error: 14.6±0.28% in humans and 12.7±0.26% in mice), ncEPs exhibit significantly higher IDR proportions (Fig. 2E), with medians of 62.5±1.42% and 60.0±1.27%, respectively. These results indicate that most ncEPs lack structured domains and are instead dominated by IDRs.

Among the eight domains observed in at least five ncORFs, three are DNA-binding domains: zinc finger, Krüppel-associated box, and helix-turn-helix (Fig. 2F). Because transposable elements (TEs) commonly encode DNA-binding proteins required for transposition49, we tested whether domain-containing ncORFs preferentially overlap TE-derived sequences. Indeed, domain-containing ncORFs were significantly enriched for TE overlap in both humans and mice (P = 5.8×10−4 and 3.4×10−17, respectively; Fisher’s exact tests; Fig. 2G). Although enrichment alone does not establish origin, it supports a model in which a subset of longer ncORFs may have acquired modular domains through TE-derived sequence contributions.

Putative ncEPs are selectively constrained

Given the scarcity of known domains in putative ncEPs, a key question is whether they are functional in cells. If at least some of these ncEPs are not merely translational noise or byproducts of translation-dependent functions, we would expect them to exhibit evolutionary signatures of selective constraints similar to canonical CDSs. To test this, we leveraged the recent high-resolution genomic constraint maps derived from both intra- and inter-species comparisons in humans 50,51. First, we examined the genomic non-coding constraint of haploinsufficient variation (Gnocchi), which quantifies selective pressure by comparing observed versus expected variants across human genomes. A Gnocchi score above zero indicates purifying selection50. Although ncORFs exhibit lower Gnocchi scores than CDSs, their score remain significantly above zero and exceed those of lncRNAs (Fig. S7), suggesting that ncORF sequences are under evolutionary constraint. A hallmark of functional coding regions is codon position-specific constraints, but Gnocchi scores lack the resolution needed to assess such fine-scale features. Therefore, we next analyzed the base-resolution conservation metric PhyloP52 calculated from multiple alignments of primate or mammalian genomes51. As expected, PhyloP scores in CDSs are substantially higher than those in flanking UTRs and display strong three-nucleotide periodicity, with the third codon position being less conserved than first and second (Fig. 3A). Although ncORFs (excluding those overlapping annotated CDSs) do not show higher overall PhyloP than flanking regions, they also exhibit significant three-nucleotide periodicity in both mammals (Fig. 3A) and primates (Fig. S8), suggesting that at least a subset of ncORFs are subject to similar evolutionary constraints as CDSs.

Evolutionary constraints of ncORFs.

(A) The mean PhyloP scores of 30 base pairs upstream and downstream of the ORF start or stop codons in mammals. Codons were delineated with bars of alternating colors by different frames, and nucleotides in untranslated regions were shown in grey. The significance of three-nucleotide periodicity was assessed by autocorrelation with a lag of three. (B) Cladograms of vertebrates with the number of ncORFs originating at each ancestral branch of humans. Triangles indicate species merged into larger clade for visual simplicity. (C) Relationship between ORF origination rates and node ages measured as node-to-tip distances. Origination rate was defined as the number of ncORFs that originated on a branch divided by the branch length. Blue lines show linear regression fits, and grey bands represent 95% prediction confidence intervals. Spearman’s correlation is indicated. (D) Distribution of ncORF PhyloCSF scores normalized by the number of codons in per ncORF. The dashed line denotes zero. CDS, coding sequence; ncORF, non-canonical open reading frame; ORF, open reading frame.

Motivated by the above results, we next sought to identify individual ncORFs under selective pressure. To this end, we determined their modes of origination33,34 and evolutionary history through comparative genomic analyses. Our analysis revealed that most ncORFs are evolutionarily young, with 63.0% (7,319) and 82.8% (13,508) ncORFs in humans and mice originating within the mammalian lineage, respectively (Figs. 3B & S9). In line with previous studies33,34, we estimate that at least 51.7% of human and 50.8% of mouse ncORFs arise de novo (Table S6). We also observed that the number of ncORFs originating at each ancestral node, when normalized by branch length, decreases with increasing node-to-tip distance (Fig. 3C), reflecting dynamic turnover and reduced long-term retention of ncORFs. To assess whether individual ncORFs are constrained for coding, we analyzed codon substitution patterns within ncORFs since origination33 with PhyloCSF53. PhyloCSF estimates the log-likelihood ratio of a given alignment under coding versus non-coding models; positive values indicate stronger similarity to conserved coding sequences. We found 13.8% (1,608) of human and 32.0% (5,281) of mouse ncORFs have positive PhyloCSF scores (Fig. 3D), and 10.9% (1,264) of human and 22.6% (3,721) of mouse ncORFs even have PhyloCSF exceeding 10 (Table S6), providing strong evidence that these sequences may encode functional ncEPs.

Evolutionary dynamics can also provide insights into the function of ncORFs, as functionally constrained ncORFs are less likely to be lost over time. To quantify the conservation of each ncORF since its origination, we developed a local branch length score (BLS) (Fig. 4A), adapting the “global” BLS metric previously used to assess uORF conservation across full phylogenies48,54. Unlike global BLS, local BLS is designed to detect lineage-specific constraint, making it better suited for identifying functional ncORFs that may be restricted to certain evolutionary branches. Using a local BLS threshold of 0.9, we identified 16.3% (1,889) of human and 24.8% (4,087) of mouse ncORFs as exhibiting lineage-specific conservation (Figs. 4B & S9). Notably, these ncORFs are significantly more likely to show positive PhyloCSF scores (Fig. 4C), reflecting the effectiveness of local BLS. In total, we identified 681 human and 1,622 mouse ncORFs that are both lineage-specifically conserved and show evidence of coding potential, making them strong candidates for future functional validation. Together, these findings suggest that a subset of ncORFs may encode functional proteins rather than translational byproducts.

Lineage-specifically conserved ncORFs.

(A) Schematic illustration for BLS calculation. (B) Cladograms of vertebrates showing the number of lineage-specifically conserved ncORFs (local BLS > 0.9) at each ancestral node for humans and mice. (C) Distribution of PhyloCSF scores per codon for lineage-specifically conserved ncORFs compared to all other ncORFs. BLS, branch length score; ncORF, non-canonical open reading frame; ORF, open reading frame.

Expression of ncORFs is shaped by their evolutionary dynamics

Gene expression is a tightly regulated process, and non-functional products are unlikely to be robustly expressed. To assess the expression of ncORFs, we examined their translation across diverse tissues and cell lines (Table S7). On average, 71.9% of human and 48.5% of mouse ncORFs are translated with reads per kilobase per million mapped reads (RPKM) ≥ 1, which is comparable to or slightly exceeds the proportions observed for CDSs (Fig. S10). To investigate the influence of ncORF evolution on their expression, we analyzed mean translation levels of ncORFs across different samples. We found that ncORFs with higher local BLS tend to be more highly translated in both species (Fig. S11A), whereas coding potential showed only a minor influence (Fig. S11B). To disentangle the effects of evolutionary constraint and gene age, we stratified ncORFs by origination nodes and examined their translation levels in relation to local BLS and PhyloCSF. Although older ncORFs are generally more highly translated, those with higher local BLS scores consistently show elevated translation levels compared to their age-matched counterparts (Figs. 5A & S12A). This suggests that lineage-specific conservation is associated with enhanced and possibly optimized translation, potentially reflecting acquired functional roles. Conversely, the relationship between coding potential and translation is more nuanced: younger ncORFs with positive PhyloCSF scores tend to be more highly translated, while the opposite trend is observed in older ncORFs (Figs. 5B & S12B). These patterns point to complex interplays between evolutionary constraint, coding potential, and translational output, likely reflecting divergent selective pressures across different ncORF age groups.

Evolutionary dynamics of ncORF expression.

(A) Distribution of mean translation levels of human ncORFs grouped by their origin nodes and further stratified by local BLS. Statistical significance was assessed using Wilcoxon rank-sum tests. ***, P < 0.001; **, P < 0.01; *, P < 0.05. (B) Similar to (A) but ncORFs are further stratified by local BLS. (C) Relationship between ncORF tissue specificity at translation and transcription levels. Differences were determined with Wilcoxon signed-rank tests. (D-E) Similar to (A) and (B) but showing the distribution of tissue specificity at the translation level. CDS, coding sequence; ncORF, non-canonical open reading frame; ORF, open reading frame.

While the above analyses revealed the overall trends in ncORF translation, they overlooked the heterogeneity across tissues and cell types. We found that ncORF translation exhibits strong tissue specificity (Fig. S13), which may stem from either transcriptional bias in ncORF-containing genes or post-transcriptional regulation specific to ncORFs. To disentangle these factors, we examined RNA levels of ncORF-containing genes across different tissues55,56. Although transcriptional enrichment in testis was evident in both humans and mice (Fig. S14), testis-biased translation was observed only in mice (Fig. S13). Moreover, ncORF translation displayed significantly higher tissue specificity than the transcription of their host genes in both species (Fig. 5C), indicating a substantial post-transcriptional component to their regulation. We next assessed how evolutionary characteristics shape tissue specificity. Older ncORFs tend to have lower tissue specificity than younger ones, suggesting they are more broadly translated (Figs. 5D & S12C). Intriguingly, coding potential exhibits opposite effects depending on evolutionary age: younger ncORFs with higher coding potential are more broadly expressed, whereas older ones show increased tissue specificity (Figs. 5E & S12D). These patterns may help to explain the lower average translation levels observed for older ncORFs across tissues. A similar trend was also evident for ncORFs with high local constraint scores in mice (Fig. S12C). Collectively, these results support a model in which ncORFs initially evolve widespread expression to establish interactions and avoid loss57, but may later become functionally specialized, resulting in more restricted expression.

Extensive co-translation between ncORFs and canonical CDSs

Given our earlier observation that many ncEPs lack modular domains and may act via interactions with canonical proteins, we next explored the landscape of co-translation between ncORFs and CDSs across different tissues or cell types as we expect that interacting proteins tend to be translated simultaneously spatiotemporally. Using weighted gene co-expression network analysis (WGCNA)58, we identified 61,513 human and 55,449 mouse ncORF-CDS co-translated pairs. As expected, ncORFs in mRNAs tend to co-translate with the main CDS of the same gene, even after excluding CDS-overlapping ones (Fig. S15), consistent with previous experimental findings28. Among non-overlapping co-translating ncORFs, only 10 in humans were dORFs; all others in both species were uORFs. Of note, co-translation does not contradict the canonical model in which uORFs repress downstream CDS translation. Co-translation reflects concurrent ribosome occupancy driven by overall initiation rates at the 5’ end, whereas repression concerns the fraction of initiating ribosomes that reach the CDS. Thus, absolute translation of both ORFs may increase without altering the relative inhibitory effect of the uORF59,60.

ncORF-CDS co-translation forms a highly modular bipartite network, in which one CDS may be linked to multiple ncORFs and vice versa (Fig. 6A-B). Network analysis revealed that hub CDS nodes—those with the most connections—in the largest modules are significantly enriched for functional gene categories (Fig. 6C-D). For instance, the largest module in humans is associated with chromatin-related proteins, while the second largest module in humans and the largest one in mice are linked to extracellular matrix organization. Notably, ancient ncORFs (those originating before the last common ancestor of mammals) are significantly more likely to co-translate with CDSs than younger ones (Fig. 6E), potentially reflecting a greater likelihood of acquiring functional interactions over evolutionary time. In summary, our analyses revealed that evolutionary dynamics of ncORFs have widespread impacts on their translation and potential function.

ncORF-CDS co-translation network.

(A) Bipartite network of ncORF-CDS co-translation in humans. (B) Bipartite network of ncORF-CDS co-translation in mice. ncORFs and CDSs are represented by different shapes and colored according to their cluster membership. Only the largest clusters were highlighted (top two in humans and top five in mice). (C) Top five enriched gene ontology terms for each of the two largest clusters in the human network shown in (A). (D) Top five enriched gene ontology terms for each of the three largest clusters in the mouse network shown in (B). (E) Proportion of ncORFs cotranslating with CDSs among ancient (pre-mammalian origin) versus younger (mammalian-specific) ncORFs. Differences were tested with Fisher’s exact tests. CDS, coding sequence; ncORF, non-canonical open reading frame; ORF, open reading frame.

Discussion

Reliable annotation of translated ncORFs is critical for understanding their functional roles and evolutionary significance. In this study, we developed a standardized pipeline for the annotation of translated ncORFs and applied it to both humans and mice due to the extensive publicly available Ribo-Seq datasets. Compared to previous efforts, our annotation framework offers several key improvements. First, we selected only high-quality libraries with sufficient read coverage and acceptable in-frame RPF rates. Our earlier work demonstrated that both library size and in-frame rates impact the reproducibility and accuracy of ncORF predictions20. This step helps mitigate false annotations arising from suboptimal data quality. Second, although many previously characterized functional ncEPs are associated with cancer-related ncORFs1, ncORFs may also have important functions under physiological conditions. Importantly, such ncORFs are more likely to be subject to evolutionary constraints. Therefore, we restricted our analysis to normal cells and tissues, allowing us to identify more biologically relevant and conserved ncORFs. Third, given the variable performance across ncORF detection methods and the low reproducibility of predictions between biological replicates20, we implemented stringent reproducibility criteria. Specifically, we retained only ncORFs detected by multiple methods and across independent datasets, enhancing the robustness of our final annotation. Finally, we applied a suite of well-validated filtering metrics from prior studies39–41 to minimize the inclusion of false positives, including those arising from translational noise61,62 or non-ribosomal contamination. As a result, our annotated ncORFs show significant enrichment for GENCODE ncORFs, especially those with MS evidence or reproducible detection across independent studies. These features collectively underscore the reliability and utility of our annotation for future functional and evolutionary studies of non-canonical translation.

However, our catalog may remain incomplete due to the uneven distribution of publicly available Ribo-seq datasets across tissues and cell types. Broader and more balanced tissue coverage will be necessary to generate a more comprehensive and less biased ncORF annotation, which would also enhance the cotranslation analysis. Additionally, our proteomic validation relies on previously compiled sets of MS-supported ORFs rather than de novo reanalysis of raw mass spectrometry datasets. Incorporation of candidate ncORFs into reference proteome databases prior to large-scale, standardized searches of public MS runs would substantially improve detection sensitivity and provide stronger evidence for their translational products45.

From the perspective of new gene evolution, ncORFs represent a transitional stage between non-coding DNA/RNA and protein-coding genes63,64. A major challenge in reclassifying ncORFs as bona fide coding genes lies in determining whether the putative proteins they encode are functionally relevant. Our analysis of ncEPs revealed that most lack recognizable protein domains and are enriched for IDRs. Therefore, we propose that many ncORFs may function in a non-autonomous manner by interacting with other proteins to exert their effects. This hypothesis aligns with prior studies showing physical interactions between ncEPs and known proteins28,29,34. Supporting this view, our WGCNA analysis uncovered widespread co-translation between ncORFs and canonical CDSs, suggesting potential interaction partners and functional contexts.

We also assessed whether ncORFs are subject to evolutionary constraints consistent with functional protein coding. Using both population-level and cross-species analyses, we found that a substantial fraction of ncORFs bear signatures of purifying selection, in line with previous studies33,59,60,65,66. Notably, their conservation profiles exhibit nucleotide-level features characteristic of canonical CDSs, supporting the interpretation that selection acts on the encoded peptide products rather than merely on translational activity per se. Furthermore, we reconstructed the evolutionary trajectories of thousands of ncORFs in both humans and mice and observed sustained coding potential since their origination. Complementing these findings, we identified a subset of ncORFs that preserve intact ORF structures across most descendant species, suggesting ongoing purifying selection against disabling mutations. These lineage-specifically conserved and potentially coding ncORFs likely contribute to clade-specific functional innovation and represent compelling candidates for future functional characterization.

We further explored how evolutionary history shapes ncORF translation across tissues or cell types. Our results indicate that older ncORFs are more broadly and highly translated and have established more extensive co-expression with canonical CDSs. Moreover, when comparing ncORFs of similar age, those that are lineage-specifically constrained consistently exhibit higher translation levels, reinforcing their likely functional importance. Interestingly, we found that younger ncORFs with strong evolutionary constraints tend to have lower tissue specificity in their translation, potentially reflecting a phase of widespread expression that facilitates functional recruitment. In contrast, older ncORFs with positive coding potential tend to display increased tissue specificity, implying functional specialization over evolutionary time. Collectively, these findings support a model in which ncORFs evolve from broadly expressed, weakly functional elements into specialized, conserved effectors—though further empirical work is needed to validate this hypothesis.

These patterns should nevertheless be interpreted in light of the heterogeneous nature of the public Ribo-seq datasets used in our analysis. The datasets differ in experimental protocols, sequencing depth, read quality, tissue and cell-type coverage, and other technical factors that can influence the sensitivity and specificity of ncORF detection and quantification. In particular, ncORFs with low or highly restricted translation may be preferentially missed in datasets with limited depth or suboptimal quality, potentially affecting estimates of translation breadth and tissue specificity. Systematic and standardized Ribo-seq profiling across matched tissues, developmental stages, and species will be required to more fully resolve the evolutionary dynamics of ncORF translation.

In conclusion, we provide a comprehensive annotation of translated ncORFs in humans and mice, offering a valuable resource for functional and evolutionary genomics. Our integrated analysis of sequence evolution and translational landscape sheds light on the evolutionary trajectory and functional integration of ncORFs. The consistency of these patterns in both species suggests that our findings may generalize to other mammals. However, realizing the full biological significance of ncORFs, particularly lineage-specific ones, will require experimental validation. Recent advances in high-throughput screening technologies19,29,32,67 offer a promising path forward. Lastly, although our study focused on humans and mice due to the availability of high-quality Ribo-Seq data, we anticipate that our annotation will serve as a training resource for machine learning models aimed at predicting translated ncORFs in non-model organisms, thus broadening our understanding of genome coding potential across species.

Methods

Quality control and preprocessing of Ribo-Seq libraries

Ribo-Seq libraries from normal human or mouse samples without genetic modifications or perturbations (Table S7) were downloaded from the European Nucleotide Archive (ENA) or the Short Read Archive (SRA). For ncORF annotation, libraries containing fewer than five million reads were excluded. Libraries derived from cancer tissues or cell lines were also removed based on sample metadata or the associated publications. Next, adapter sequences and any random nucleotide additions were trimmed using Cutadapt (v3.7)68 and reads shorter than 18 nucleotides were discarded. Remaining reads were aligned to the human or mouse reference genome (GRCh38 or GRCm39, ENSEMBL69 release 107) using the STAR aligner (v2.7.10a)70. To assess library quality, the proportion of in-frame ribosome-protected fragment (RPF) reads was calculated using the PSite package71. Libraries with ≥60% in-frame RPFs were classified as high quality. Only samples containing at least two high-quality libraries were retained for ncORF annotation (Table S1). For the expression analyses, all the downloaded Ribo-Seq libraries were used to cover more samples.

ncORF annotation

For each Ribo-Seq library, translated ORFs initiating with AUG, CUG, GUG, or UUG and encoding at least five amino acids were predicted using PRICE36, RiboCode37, and Ribo-TISH38, as described previously20. To define novel ORFs, annotated CDSs and their truncation or extension variants (i.e., ORFs with the same start or stop position as annotated CDSs) were excluded. To eliminate likely false positives resulting from non-ribosomal contamination or stochastic molecular errors, we calculated RRS40,41 and FLOSS39 of each ncORF. RRS measures the drop in RPF coverage after stop codon—a key signature of active translation beyond three-nucleotide periodicity40,41, whereas FLOSS assesses whether the RPF length distribution resembles that of canonical CDSs, helping to distinguish authentic ORFs from non-ribosomal signals39. In each library, uniquely mapped RPFs were assigned to the corresponding P-site, and genome-wide P-site coverage was computed with the PSite package. Similar to previous studies40,41, RRS was calculated as the sum of coverage across the five codons preceding the stop codon divided by the sum across the five trinucleotides following the stop codon. FLOSS was calculated by first deriving a reference RPF length distribution vector (Pref) using all RPFs uniquely mapped to annotated CDSs. For each ORF (ncORF or CDS), a read length frequency vector (PORF) was computed, and FLOSS was calculated as ∑|Pref − PORF|/2. The empirical upper bound for acceptable FLOSS was set using Tukey’s method (i.e., the third quartile + 1.5 × inter-quartile range of FLOSS values across all CDSs)39. ncORFs with fewer than 10 P-site-mapped RPFs, RRS ≤ 1, or FLOSS values exceeding the empirical upper bound were filtered out.

Integration of ncORF prediction across different libraries of the same species was nontrivial due to extensive overlap among ncORFs in genomic coordinates. To address this issue, we used a graph-based clustering approach. We created a graph where two ORF vertices were connected if they share the same start or stop codon. Then, each set of connected ncORFs in the graph represents a cluster of overlapping ncORFs. To ensure reproducibility, only clusters identified by ≥2 methods in ≥2 libraries were retained, yielding 13,071 clusters in humans and 17,674 clusters in mice. To ensure compatibility with the latest reference gene annotations (ENSEMBL release 111), we removed any ncORFs that were newly classified as annotated CDSs. A manual inspection of RPF signals using the GWIPS-viz Ribo-Seq track72 in the UCSC Genome Browser73 revealed additional spurious ncORFs overlapping in-frame with CDSs from other transcripts. These were likely misidentified due to shared translation signals and were conservatively excluded using custom scripts, leaving 11,623 clusters in humans and 16,485 in mice.

To facilitate downstream analyses, a representative ncORF was selected for each cluster using a two-step strategy. First, a representative transcript was selected by prioritizing those with matched annotation from NCBI and EMBL-EBI (MANE)42 (for humans) or APPRIS principal74; otherwise, the longest transcript was chosen. Then, among ncORFs within the representative transcript, a single ncORF was selected based on start codon hierarchy (AUG > CUG/GUG > UUG) following empirical evidence28,36. If multiple ncORFs shared the same preferred start codon, the longer one was chosen. Finally, we classified ncORFs according to their positions relative to annotated transcript features. ncORFs located in lncRNAs were designated as lncORFs. Among ncORFs in protein-coding transcripts, those located entirely within the 5′ UTR were classified as uORFs, whereas those initiating in the 5′ UTR and extending into the CDS were classified as uoORFs. ncORFs with both start and stop codons within the annotated CDS were classified as iORFs, whereas those initiating within the CDS and terminating in the 3′ UTR were classified as doORFs. Those initiating in the 3′ UTR were classified as dORFs.

For validation, our ncORFs were compared with previously annotated GENCODE ncORFs2 and those with MS evidence compiled previously20. Tier annotations of GENCODE ncORFs were obtained from a recent study45. Known functional ncORFs and their genomic coordinates were manually curated from published literature (Table S4). To account for possible proteoforms arising from alternative initiation and splicing, we considered an ncORF to be recovered if either its start or stop codon was identical to any ncORF in our annotation.

ncORF sequence analyses

The secondary structure surrounding each ORF (ncORF or CDS) start codon was predicted with RNAfold75 on a 150 nt window extracted from the transcript in which the ncORF resides starting from the 30th nucleotide upstream of the start codon. For each position within this window, the overall pairing probability was calculated by dividing the number of ORFs with a base-paired residue at that position by the total number of ORFs analyzed. The tRNA adaptation index and effective number of codons for ncORFs and CDSs were calculated using the “cubar” R package76. tRNA gene annotations in humans and mice were obtained from GtRNAdb (release 22)77.

Putative ncEPs analyses

Known domains within ncEPs were identified using InterProScan (v5.64-96.0)78 with default parameters. Repetitive sequence annotations were downloaded from the UCSC Table Browser. BEDTools (v2.31.1)79 was used to identify TE-overlapping ncORFs. IDRs were predicted with IUPred380 with analysis type set to “short disorders” and smoothing disabled. For each ORF, residues with disorder tendency score > 0.5 were considered disordered, and the IDR fraction was calculated as the proportion of disordered residues relative to the ORF length.

Gnocchi Score and PhyloP of ncORFs and CDSs

Genome-wide Gnocchi scores computed using 1 kb sliding windows (with 100bp step size) were downloaded from the gnomAD database50. For each base pair, the Gnocchi score was defined as the average across the ten overlapping 1kb windows that covered that position using BEDTools. The final Gnocchi score for each ORF (ncORF or CDS) was obtained by averaging the scores of all base pairs within the ORF.

Base-wise PhyloP scores computed across 441 mammals or 239 primates51 were downloaded from the UCSC Genome Browser. For each ncORF or CDS, PhyloP values were extracted for a ±30 bp window surrounding the start and stop codons using custom scripts. These values were then averaged over different ORFs to generate metagene profiles for visualization. To evaluate three-nucleotide periodicity, the metagene profiles were subjected to autocorrelation analysis using the “ac.test” function from the “testcorr” R package81. Statistical significance was assessed using the robust t̃ statistics.

ncORF origination node and mode

Whole genome alignments and corresponding species trees were downloaded from the UCSC Genome Browser (GRCh38 100-way for humans and GRCm39 35-way for mice). The origination node and mode for each ncORF were inferred following previous approaches33,34. For each ORF, the sequence from the reference genome and its orthologous regions in other species were extracted. Orthologous sequences covering less than 50% of the ncORF length relative to the reference species were excluded. The remaining sequences were realigned with MAFFT (v7.520)82 with default parameters. The corresponding gene tree was derived by pruning the species tree using Gotree (v0.4.3)83 with default settings. Ancestral sequence reconstruction was performed using FastML (v3.11)84 with parameters “--seqType nuc --SubMatrix HKY --OptimizeBL no --indelReconstruction ML”. An ORF structure was considered intact in an orthologous or ancestral sequence if one of the first three codons was a start codon and at least 70% of the 5’ region of the reference ncORF was free of in-frame stop codons. The origination node was defined as the most ancient ancestral node that retained an intact ORF structure. An ncORF was classified as de novo if the origination node had an ancestor covering at least 50% of the ncORF sequence but lacking an intact ncORF structure. Otherwise, the origination mode was classified as uncertain. The distance from an ancestral node to the reference species (tip node) was calculated by summing branch lengths along the connecting path using the ape R package (v5.8)85.

ncORF BLS and PhyloCSF

For each ncORF, global BLS was calculated as the ratio of the total branch length of the subtree connecting all leaves with intact ORF structures to the total branch length of the full species tree as in our previous study48 with custom scripts. Local BLS was calculated similarly, but the denominator was replaced by the total branch length of the subtree defined by the origination node and all of its descendant nodes (including leaves). To evaluate coding potential, the alignment of all species descending from the origination node was converted into Multiple Alignment Format after removing the stop codon. PhyloCSF scores were then computed using the “score-msa” module of PhyloCSF++86.

ncORF expression analysis

To quantify translation levels of ncORFs and CDSs in Ribo-Seq libraries, we assigned uniquely mapped reads to corresponding P-site positions with the PSite package. The read count for each ORF was calculated as the total number of RPFs whose P-sites fell within the ORF. To reduce bias caused by RPF enrichment at ORF termini, reads mapping to start or stop codons were excluded. For CDSs, only the count of the representative transcript of each gene as defined above was used for downstream analyses. For samples with multiple libraries, read counts were aggregated prior to normalization. Sample-specific size factors were estimated using CDS counts and applied to normalize both ncORF and CDS counts across samples using DESeq287. RPKM for an ORF was calculated as normalized read counts / ORF length (nt) / library size × 109, where library size was the median of total read counts across samples to ensure cross-sample comparability. An ORF (ncORF or CDS) was considered translated if its RPKM was ≥ 1. To evaluate tissue specificity of ncORF translation, the normalized RPF count Xk of an ORF in sample k was log-transformed as xk = log(Xk + 1). Then, tissue specificity index88 was calculated as τ=∑kn(1−x~k)/(n−1), where x~k=xk/max1≤j≤nxj and n is the total number of samples. To quantify RNA levels, raw reads of deep transcriptome profiling across different human or mouse tissues from previous studies55,56 were downloaded from SRA. Gene expression levels (transcript per million, TPM) were estimated using salmon (v1.10.3)89 with parameters “--seqBias --gcBias --posBias -l A”.

WGCNA analysis58 of ncORF and CDS translation was performed using the PyWGCNA package (v2.0.5)90 with default settings. Only ORFs with normalized RPF counts ≥ 10 in more than half of the samples were included. Pairwise cotranslation relationships were quantified using the topological overlap measure (TOM) derived from the resulting networks. Because the distributions of TOM values differed between the human and mouse networks following species-specific soft-threshold selection, empirical TOM cutoffs were determined separately for each species. Two ORFs were considered cotranslated if their TOM exceeded 0.1 (human) or 0.01 (mouse).58 To test whether co-translation is enriched between ncORFs and CDSs from the same transcript, gene identities of CDSs were randomly permuted among all co-translated ncORF–CDS pairs. This procedure was repeated 1,000 times, and the empirical P value was calculated as (m + 1) / 1,000, where m is the number of permutations in which the number of same-gene pairs was greater than or equal to the observed count. The bipartite network of ncORF-CDS co-translation was clustered using the Leiden algorithm91 implemented in igraph (v2.1.3)92 by optimizing modularity with a resolution of 0.2. The top 20% of nodes by betweenness centrality were designated as hub nodes. Gene ontology analysis of hub genes in the largest modules was conducted using ClusterProfiler (v4.12)93.